---
title: "02. IPA Analysis"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document:
    latex_engine: xelatex
    keep_tex: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## I. Set Up 

### A. Load Libraries

```{r warning =FALSE, message =FALSE}
library(data.table)
library(DataExplorer)
library(fastLink)
library(furniture)
library(ggrepel)
library(ggthemes)
library(haven)
library(here)
library(Hmisc)
library(infer)
library(ipumsr)
library(kableExtra)
library(knitr)
library(latex2exp)
library(lubridate)
library(matrixStats)
library(modeest)
library(pastecs)
library(psych)
library(purrr)
library(readxl)
library(scales)
library(stargazer)
library(stringr)
library(survey)
library(tibble)
library(tidyverse)
library(writexl)
library(srvyr)

```

### B. Clear Memory
```{r}
rm(list = ls(all.names = TRUE))
gc()
```

### C. Custom Functions

```{r}
#some self-created functions that I frequently use
freq <- function(x) {
frequency <- table(x, useNA = "ifany")
proportion <- round(prop.table(table(x, useNA = "ifany"),),3)
cbind(frequency, proportion)
}

freq_sort <- function(x) {
frequency <- table(x, useNA = "ifany")
proportion <- round(prop.table(table(x, useNA = "ifany"),),3)
out <- cbind(frequency, proportion)
out[order(out[,1],decreasing=TRUE),]
}

freq2 = function(x,y){
table(x,y, useNA = "ifany")
}

freq2_col <-function(x,y){
prop.table(table(x,y,useNA = "ifany"),2)
}

freq2_row <-function(x,y){
prop.table(table(x,y,useNA = "ifany"),1)
}

  ttable <- function(dependent, by_variable){
  level <- c("By variable=0", "By variable=1", "Difference")
  means <- round(c(mean(dependent[by_variable==0], na.rm=TRUE),
  mean(dependent[by_variable==1], na.rm=TRUE),
  mean(dependent[by_variable==1], na.rm=TRUE) -
  mean(dependent[by_variable==0], na.rm=TRUE)),3)
  se_0 <- sd(dependent[by_variable==0],
  na.rm=TRUE)/sqrt(sum(!is.na(dependent[by_variable==0])))
  se_1 <- sd(dependent[by_variable==1],
  na.rm=TRUE)/sqrt(sum(!is.na(dependent[by_variable==1])))
  se_diff <-sqrt(se_0^2+ se_1^2)
  se <- round(c(se_0, se_1, se_diff),3)
  data_correction <- c(1,1,2)
  N <- c(sum(!is.na(dependent[by_variable==0])) ,
  sum(!is.na(dependent[by_variable==01])), sum(!is.na(dependent)))
  results <- as.data.frame(cbind(level, means, se, N, data_correction))
  results$means <-as.numeric(results$means)
  results$se <-as.numeric(results$se)
  results$N <-as.numeric(results$N)
  results$data_correction <-as.numeric(results$data_correction)
  ttable <- results %>%
  mutate(t_stat = round(means/se,3),
  data_t = N - data_correction,
  p_value = 2*pt(abs(t_stat), data=data_t, lower.tail=FALSE)
  ) %>%
  select(level, means, se, t_stat, data_t, p_value )
  rm(level, means, se_0, se_1, se_diff, se, data_correction, N, results)
  ttable
}

```

## II. Data Loading

### A. Load IPUMS data 
```{r}
data <- readRDS("../data/clean/01_IPA_cleaning_household_clean.rds")


print(object.size(data), units = "GB")
```


```{r}
plot_missing(data)
```


### C.Create survey design objects 

About the survey: 
- To make accurate inferences about the full California population, we must apply survey weights and account for the sampling design. 
- This includes adjusting for clustering (primary sampling units) and stratification,  which ensures that standard errors and confidence intervals are correctly estimated.
- Because this is not a simple random sample, failing to account for design effects would lead to biased estimates and misleading precision.
- svy_households <- svydesign(ids = ~cluster,strata = ~strata,weights = ~weight_household,data = data, nest = TRUE): 
  - id = cluster:	Accounts for clustering in the sample design (e.g., multiple people from same tract)
  - strata = ~strata: Ensures stratified sampling is accounted for (e.g., sampling by region or income bin)
  - weights = ~weight_household	:Makes the estimates representative of the full population


```{r}
# Join AMI + define key variables for survey analysis
data <- data %>%
  mutate(
    # Ensure types
    income = as.numeric(income),
    weight_household_norm = as.numeric(weight_household_norm),
    multyear = as.integer(multyear),

    # Demographic dummies
    white_head = race_ethnicity_head == "White",
    black_head = race_ethnicity_head == "Black",
    latinx_head = str_detect(race_ethnicity_head, "Hispanic"),
    elderly_head = age_head_group %in% c("65-74", "75+"),

    # Income categories (HCD-style)
    income_category = case_when(
      percent_ami < 0.30 ~ "Extremely Low Income",
      percent_ami < 0.50 ~ "Very Low Income",
      percent_ami < 0.80 ~ "Low Income",
      percent_ami < 1.20 ~ "Moderate Income",
      percent_ami >= 1.20 ~ "Above Moderate Income",
      TRUE ~ NA_character_
    ),
    income_category = factor(
      income_category,
      levels = c(
        "Extremely Low Income", "Very Low Income", "Low Income",
        "Moderate Income", "Above Moderate Income"
      )
    ),
    low_income = income_category %in% c(
      "Extremely Low Income", "Very Low Income", "Low Income"
    ),
    moderate_income = income_category == "Moderate Income",

    # Likely eligible for first-time homebuyer (FTB) assistance program
    likely_ftb_eligible = case_when(
      is_renter == 1 & income <= 1.2 * ami ~ 1,
      TRUE ~ 0
    )
  )

# Estimate averages using survey design
# - This creates a survey design object using the {survey} package. It tells R how to account for:
#     - Clusters (e.g., PSU)
#     - Strata (e.g., sampling strata)
#     - Weights (e.g., normalized household weights)

# This object (svy_households) is the foundational structure that contains both data and the metadata needed for valid variance estimation, confidence intervals, and standard errors using Taylor series linearization.
svy_households <- svydesign(
  ids = ~cluster,
  strata = ~strata,
  weights = ~weight_household_norm,
  data = data,
  #nest = TRUE
)

# This converts the design object to a tbl_svy object using the {srvyr} package. It lets you work more naturally with tidyverse workflows
srvy <- as_survey(svy_households)

```


```{r tenure}
# A. Tenure
# year 
tenure_by_year_tenure_svy <- srvy %>%
  group_by(multyear) %>%
  summarize(
    owner_mean = survey_mean(is_owner, vartype = "se", na.rm = TRUE),
    renter_mean = survey_mean(is_renter, vartype = "se", na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(owner_mean, renter_mean),
    names_to = "tenure_detailed",
    values_to = "share"
  ) %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "owner_mean" ~ "Owner",
      tenure_detailed == "renter_mean" ~ "Renter",
      TRUE ~ tenure_detailed
    )
  )

# Veteran (REVISE)
tenure_veteran_svy <- srvy %>%
  group_by(veteran_head) %>%
  summarize(share = survey_mean(is_owner, vartype = c("se", "ci")))

# Disability  
tenure_disability_svy <- srvy %>%
  group_by(disability_head) %>%
  summarize(across(
    c(is_owner, is_renter),
    ~ survey_mean(.x, vartype = "se", na.rm = TRUE),
    .names = "{.col}"
  )) %>%
  pivot_longer(
    cols = starts_with("is_"),
    names_to = "tenure_detailed",
    values_to = "share"
  ) %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "is_owner" ~ "Owner",
      tenure_detailed == "is_renter" ~ "Renter",
      TRUE ~ tenure_detailed
    )
  )

# gender only
tenure_gender_svy <- srvy %>%
  group_by(female_head) %>%
  summarize(
    owner_mean = survey_mean(is_owner, vartype = "se", na.rm = TRUE),
    renter_mean = survey_mean(is_renter, vartype = "se", na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(owner_mean, renter_mean),
    names_to = "tenure_detailed",
    values_to = "share"
  ) %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "owner_mean" ~ "Owner",
      tenure_detailed == "renter_mean" ~ "Renter",
      TRUE ~ tenure_detailed
    )
  )

# gender and year
tenure_by_gender_year_svy <- srvy %>%
  group_by(multyear, female_head) %>%
  summarize(
    owner_mean = survey_mean(is_owner, vartype = "se", na.rm = TRUE),
    renter_mean = survey_mean(is_renter, vartype = "se", na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(owner_mean, renter_mean),
    names_to = "tenure_detailed",
    values_to = "share"
  ) %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "owner_mean" ~ "Owner",
      tenure_detailed == "renter_mean" ~ "Renter",
      TRUE ~ tenure_detailed
    )
  )



# age head 
tenure_by_age_svy <- srvy %>%
  group_by(age_head_group) %>%
  summarize(
    owner_mean = survey_mean(is_owner, vartype = "se", na.rm = TRUE),
    renter_mean = survey_mean(is_renter, vartype = "se", na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(owner_mean, renter_mean),
    names_to = "tenure_detailed",
    values_to = "share"
  ) %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "owner_mean" ~ "Owner",
      tenure_detailed == "renter_mean" ~ "Renter",
      TRUE ~ tenure_detailed
    )
  )

# age head and year 
tenure_by_age_year_svy <- srvy %>%
  group_by(multyear, age_head_group) %>%
  summarize(
    owner_mean = survey_mean(is_owner, vartype = "se", na.rm = TRUE),
    renter_mean = survey_mean(is_renter, vartype = "se", na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(owner_mean, renter_mean),
    names_to = "tenure_detailed",
    values_to = "share"
  ) %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "owner_mean" ~ "Owner",
      tenure_detailed == "renter_mean" ~ "Renter",
      TRUE ~ tenure_detailed
    )
  )

# age, year, female 
tenure_by_age_year_svy <- srvy %>%
  group_by(multyear, age_head_group, female_head) %>%
  summarize(
    owner_mean = survey_mean(is_owner, vartype = "se", na.rm = TRUE),
    renter_mean = survey_mean(is_renter, vartype = "se", na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(owner_mean, renter_mean),
    names_to = "tenure_detailed",
    values_to = "share"
  ) %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "owner_mean" ~ "Owner",
      tenure_detailed == "renter_mean" ~ "Renter",
      TRUE ~ tenure_detailed
    )
  )


#race 
tenure_by_race_svy <- srvy %>%
  group_by(race_ethnicity_head) %>%
  summarize(
    owner_mean = survey_mean(is_owner, vartype = "se", na.rm = TRUE),
    renter_mean = survey_mean(is_renter, vartype = "se", na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(owner_mean, renter_mean),
    names_to = "tenure_detailed",
    values_to = "share"
  ) %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "owner_mean" ~ "Owner",
      tenure_detailed == "renter_mean" ~ "Renter",
      TRUE ~ tenure_detailed
    )
  )

# race + year 
tenure_by_race_svy <- srvy %>%
  group_by( multyear,race_ethnicity_head) %>%
  summarize(
    owner_mean = survey_mean(is_owner, vartype = "se", na.rm = TRUE),
    renter_mean = survey_mean(is_renter, vartype = "se", na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(owner_mean, renter_mean),
    names_to = "tenure_detailed",
    values_to = "share"
  ) %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "owner_mean" ~ "Owner",
      tenure_detailed == "renter_mean" ~ "Renter",
      TRUE ~ tenure_detailed
    )
  )

# race + year + female 
tenure_by_race_svy <- srvy %>%
  group_by( multyear,race_ethnicity_head, female_head) %>%
  summarize(
    owner_mean = survey_mean(is_owner, vartype = "se", na.rm = TRUE),
    renter_mean = survey_mean(is_renter, vartype = "se", na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(owner_mean, renter_mean),
    names_to = "tenure_detailed",
    values_to = "share"
  ) %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "owner_mean" ~ "Owner",
      tenure_detailed == "renter_mean" ~ "Renter",
      TRUE ~ tenure_detailed
    )
  )

# income 
ownership_by_income_svy <- srvy %>%
  group_by(income_bracket) %>%
  summarize(
    owner_mean = survey_mean(is_owner, vartype = "se", na.rm = TRUE),
    renter_mean = survey_mean(is_renter, vartype = "se", na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(owner_mean, renter_mean),
    names_to = "tenure_detailed",
    values_to = "share"
  ) %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "owner_mean" ~ "Owner",
      tenure_detailed == "renter_mean" ~ "Renter",
      TRUE ~ tenure_detailed
    )
  )

# Tenure by Inccome + year 
tenure_by_income_year_svy <- srvy %>%
  filter(!is.na(income_bracket)) %>%
  group_by(multyear, income_bracket) %>%
  summarize(
    owner_mean = survey_mean(is_owner, vartype = "se", na.rm = TRUE),
    renter_mean = survey_mean(is_renter, vartype = "se", na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(owner_mean, renter_mean),
               names_to = "tenure_detailed",
               values_to = "share") %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "owner_mean" ~ "Owner",
      tenure_detailed == "renter_mean" ~ "Renter"
    )
  )

# Tenure by Ami + year 
tenure_by_ami_year_svy <- srvy %>%
  filter(!is.na(ami_category)) %>%
  group_by(multyear, ami_category) %>%
  summarize(
    owner_mean = survey_mean(is_owner, vartype = "se", na.rm = TRUE),
    renter_mean = survey_mean(is_renter, vartype = "se", na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(owner_mean, renter_mean),
               names_to = "tenure_detailed",
               values_to = "share") %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "owner_mean" ~ "Owner",
      tenure_detailed == "renter_mean" ~ "Renter"
    )
  )


# ami 
ownership_by_ami_svy <- srvy %>%
  group_by(ami_category) %>%
  summarize(
    owner_mean = survey_mean(is_owner, vartype = "se", na.rm = TRUE),
    renter_mean = survey_mean(is_renter, vartype = "se", na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(owner_mean, renter_mean),
    names_to = "tenure_detailed",
    values_to = "share"
  ) %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "owner_mean" ~ "Owner",
      tenure_detailed == "renter_mean" ~ "Renter",
      TRUE ~ tenure_detailed
    )
  )

# cost burden
ownership_by_cost_burden_svy <- srvy %>%
  group_by(cost_burden_category) %>%
  summarize(
    owner_mean = survey_mean(is_owner, vartype = "se", na.rm = TRUE),
    renter_mean = survey_mean(is_renter, vartype = "se", na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(owner_mean, renter_mean),
    names_to = "tenure_detailed",
    values_to = "share"
  ) %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "owner_mean" ~ "Owner",
      tenure_detailed == "renter_mean" ~ "Renter",
      TRUE ~ tenure_detailed
    )
  )

#  tenure by cost burden + year 
tenure_by_cost_burden_year_svy <- srvy %>%
  filter(!is.na(cost_burden_category)) %>%
  group_by(multyear, cost_burden_category) %>%
  summarize(
    owner_mean = survey_mean(is_owner, vartype = "se", na.rm = TRUE),
    renter_mean = survey_mean(is_renter, vartype = "se", na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(owner_mean, renter_mean),
               names_to = "tenure_detailed",
               values_to = "share") %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "owner_mean" ~ "Owner",
      tenure_detailed == "renter_mean" ~ "Renter"
    )
  )

# tenure , income, race
# disparities in ownership rates across income levels and racial groups
tenure_by_income_race_svy <- srvy %>%
  filter(!is.na(income_bracket), !is.na(race_ethnicity_head)) %>%
  group_by(income_bracket, race_ethnicity_head) %>%
  summarize(
    owner_mean = survey_mean(is_owner, vartype = "se", na.rm = TRUE),
    renter_mean = survey_mean(is_renter, vartype = "se", na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(owner_mean, renter_mean),
               names_to = "tenure_detailed",
               values_to = "share") %>%
  mutate(
    tenure_detailed = case_when(
      tenure_detailed == "owner_mean" ~ "Owner",
      tenure_detailed == "renter_mean" ~ "Renter"
    )
  )
```

```{r}
# Cost Burden
# Cost Buurden by income bracket
cost_burden_by_income_svy <- srvy %>%
  filter(!is.na(cost_burden_category), !is.na(income_bracket)) %>%
  group_by(income_bracket, cost_burden_category) %>%
  summarize(share = survey_mean(overburdened, vartype = "se", na.rm = TRUE))

# Cost Burden by Tenure 
cost_burden_by_tenure_svy <- srvy %>%
  filter(!is.na(cost_burden_category)) %>%
  group_by(tenure_general, cost_burden_category) %>%
  summarize(share = survey_mean(overburdened, vartype = "se", na.rm = TRUE))

# Cost burden, income , race 
# --> This shows whether households with the same income still experience different cost burdens by race, which can reflect structural inequalities like discrimination, neighborhood sorting, or wealth gaps.
cost_burden_by_income_race_svy <- srvy %>%
  filter(!is.na(cost_burden_category), !is.na(income_bracket), !is.na(race_ethnicity_head)) %>%
  group_by(income_bracket, race_ethnicity_head, cost_burden_category) %>%
  summarize(share = survey_mean(overburdened, vartype = "se", na.rm = TRUE))

# cost burden  ami, race/ethnicity
# --> affordability context to racial disparities in cost burden.
cost_burden_by_ami_race_svy <- srvy %>%
  filter(!is.na(cost_burden_category), !is.na(ami_category), !is.na(race_ethnicity_head)) %>%
  group_by(ami_category, race_ethnicity_head, cost_burden_category) %>%
  summarize(share = survey_mean(overburdened, vartype = "se", na.rm = TRUE))

```

```{r}
# Mortgage
# Mortgage Burden by AMI
mortgage_burden_by_ami_svy <- srvy %>%
  filter(!is.na(mortgage_burden), !is.na(ami_category)) %>%
  group_by(ami_category) %>%
  summarize( avg_burden = survey_mean(mortgage_burden, vartype = "se", na.rm = TRUE))
    
# mortgage burden, ami, gender 
mortgage_burden_by_ami_gender_svy <- srvy %>%
  filter(!is.na(mortgage_burden), !is.na(ami_category)) %>%
  group_by(ami_category, female_head) %>%
  summarize(avg_burden = survey_mean(mortgage_burden, vartype = "se", na.rm = TRUE))

    
  # Tenure by mortgage burden 
tenure_by_mortgage_burden_svy <- srvy %>%
  filter(is_owner == 1, !is.na(mortgage_burden)) %>%
  mutate(
    mortgage_burden_category = case_when(
      mortgage_burden < 0.3 ~ "<30%",
      mortgage_burden < 0.5 ~ "30–49%",
      mortgage_burden >= 0.5 ~ "50%+",
      TRUE ~ NA_character_)) %>%
  group_by(mortgage_burden_category) %>%
  summarize(share = survey_mean(is_owner, vartype = "se", na.rm = TRUE))
```

#### TO BE UPDATED
Save precomputed survey summaries
Note: We are not going to reevaluate them after this original run since they take a long time, so we are saving them and then reload them later when we need it 
```{r save_survey_outputs, message=FALSE, eval=False}
# ---------------------------
# A. Tenure trends over time
# ---------------------------
saveRDS(ownership_year_svy, "../data/clean/svys/ownership_year_svy.rds")
saveRDS(rent_year_svy,"../data/clean/svys/rent_year_svy.rds")
saveRDS(ownership_by_year_tenure_svy,"../data/clean/svys/ownership_by_year_tenure_svy.rds")

# ---------------------------
# B. Tenure by demographics
# ---------------------------
saveRDS(tenure_veteran_svy,"../data/clean/svys/tenure_veteran_svy.rds")
saveRDS(tenure_disability_svy,"../data/clean/svys/tenure_disability_svy.rds")
saveRDS(tenure_age_svy,"../data/clean/svys/tenure_age_svy.rds")
saveRDS(ownership_gender_svy,"../data/clean/svys/ownership_gender_svy.rds")
saveRDS(ownership_by_race_svy,"../data/clean/svys/ownership_by_race_svy.rds")



# ---------------------------
# C. Tenure by race/ethnicity
# ---------------------------
saveRDS(ownership_race_svy,"../data/clean/svys/ownership_race_svy.rds")

# ---------------------------
# D. Tenure by income & affordability
# ---------------------------
saveRDS(ownership_income_svy,"../data/clean/svys/ownership_income_svy.rds")
saveRDS(ownership_ami_svy,"../data/clean/svys/ownership_ami_svy.rds")

# ---------------------------
# E. Cost burden
# ---------------------------
saveRDS(burden_race_svy,"../data/clean/svys/burden_race_svy.rds")
saveRDS(severe_burden_race_svy,"../data/clean/svys/severe_burden_race_svy.rds")

# ---------------------------
# F. Geography
# ---------------------------
saveRDS(ownership_puma_svy,"../data/clean/svys/ownership_puma_svy.rds")

# ---------------------------
# G. Analysis flags
# ---------------------------
saveRDS(farmworker_homeownership_svy,"../data/clean/svys/farmworker_homeownership_svy.rds")
saveRDS(ftb_eligibility_svy,"../data/clean/svys/ftb_eligibility_svy.rds")

```

```{r, eval =False}
# ---------------------------
# A. Tenure trends over time
# ---------------------------
ownership_year_svy <- readRDS("../data/clean/svys/ownership_year_svy.rds")
rent_year_svy <- readRDS("../data/clean/svys/rent_year_svy.rds")
ownership_by_year_tenure_svy <- readRDS("../data/clean/svys/ownership_by_year_tenure_svy.rds")

# ---------------------------
# B. Tenure by demographics
# ---------------------------
tenure_veteran_svy <- readRDS("../data/clean/svys/tenure_veteran_svy.rds")
tenure_disability_svy <- readRDS("../data/clean/svys/tenure_disability_svy.rds")
tenure_age_svy <- readRDS("../data/clean/svys/tenure_age_svy.rds")
ownership_gender_svy <- readRDS("../data/clean/svys/ownership_gender_svy.rds")

# ---------------------------
# C. Tenure by race/ethnicity
# ---------------------------
ownership_race_svy <- readRDS("../data/clean/svys/ownership_race_svy.rds")

# ---------------------------
# D. Tenure by income & affordability
# ---------------------------
ownership_income_svy <- readRDS("../data/clean/svys/ownership_income_svy.rds")
ownership_ami_svy <- readRDS("../data/clean/svys/ownership_ami_svy.rds")

# ---------------------------
# E. Cost burden
# ---------------------------
burden_race_svy <- readRDS("../data/clean/svys/burden_race_svy.rds")
severe_burden_race_svy <- readRDS("../data/clean/svys/severe_burden_race_svy.rds")

# ---------------------------
# F. Geography
# ---------------------------
ownership_puma_svy <- readRDS("../data/clean/svys/ownership_puma_svy.rds")

# ---------------------------
# G. Analysis flags
# ---------------------------
farmworker_homeownership_svy <- readRDS("../data/clean/svys/farmworker_homeownership_svy.rds")
ftb_eligibility_svy <- readRDS("../data/clean/svys/ftb_eligibility_svy.rds")
```

## III. Homeownership Trends (2018–2024)

### A. Ownership Rate by Year and Tenure 
```{r ownership_year_new}

ggplot(tenure_by_year_tenure_svy, aes(x = factor(multyear), y = share, fill = tenure_detailed)) +
  geom_col(position = "fill", width = 0.7) +  # Use "fill" for stacked percentages
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Homeownership and Rentership Rates in California (2014–2023)",
    subtitle = "Based on survey-weighted estimates, by household tenure",
    x = "Year",
    y = "Share of Households",
    fill = "Tenure"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )


```

### B. Homeownership over time with confidence intervals (Revise)
```{r ownership_year_new}
tenure_by_year_tenure_svy %>%
  filter(tenure_detailed == "Owner") %>%
  ggplot(aes(x = factor(multyear), y = share)) +
  geom_col(fill = "steelblue", width = 0.7) +
  geom_errorbar(aes(
    ymin = share - owner_mean_se,
    ymax = share + owner_mean_se
  ), width = 0.2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    title = "Survey-Weighted Homeownership Rates Over Time (with SE)",
    x = "Year", y = "Homeownership Rate"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


```



## IV. Demographic Disparities
### A. By Race/Ethnicity (KP.REVIST)
```{r race_ethnicity_w}
tenure_by_race_svy %>%
  filter(tenure_detailed == "Owner") %>%
  ggplot(aes(x = reorder(race_ethnicity_head, share), y = share, fill = race_ethnicity_head)) +
  geom_col(show.legend = FALSE, width = 0.7) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    title = "Survey-Weighted Homeownership by Race/Ethnicity",
    x = "Race/Ethnicity of Household Head",
    y = "Homeownership Rate"
  ) +
  coord_flip() +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )



```

### B. By Income Category
```{r income_group_w}
tenure_by_income_year_svy %>%
  filter(tenure_detailed == "Owner") %>%
  ggplot(aes(x = multyear, y = share, color = income_bracket)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    title = "Trends in Homeownership by Income Bracket (2014–2023)",
    x = "Year", y = "Homeownership Rate",
    color = "Income Bracket"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

```

### C.
```{r }
tenure_by_income_year_svy %>%
  filter(tenure_detailed == "Owner") %>%
  ggplot(aes(x = factor(multyear), y = share)) +
  geom_col(fill = "steelblue", width = 0.7) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~income_bracket) +
  labs(
    title = "Homeownership Rates by Income Bracket and Year",
    x = "Year", y = "Homeownership Rate"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )


```

```{r}
tenure_by_income_year_svy %>%
  filter(tenure_detailed == "Owner") %>%
  ggplot(aes(x = income_bracket, y = share, fill = income_bracket)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~multyear, ncol = 5) +
  labs(
    title = "Homeownership by Income Bracket Across Years",
    x = "Income Bracket", y = "Homeownership Rate"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )

```
### D.  Disability Status
```{r Disability}
# Filter to drop SE rows and reshape the SEs into a column
ownership_disability_clean <- tenure_disability_svy %>%
  filter(!str_detect(tenure_detailed, "_se")) %>%
  left_join(
   tenure_disability_svy%>%
      filter(str_detect(tenure_detailed, "_se")) %>%
      mutate(tenure_detailed = str_remove(tenure_detailed, "_se")) %>%
      rename(share_se = share),
    by = c("disability_head", "tenure_detailed")
  )

ggplot(ownership_disability_clean, aes(x = tenure_detailed, y = share, fill = disability_head)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = share - share_se, ymax = share + share_se),
                position = position_dodge(width = 0.8), width = 0.2) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Survey-Weighted Tenure by Disability Status",
    x = "Tenure Type", y = "Share of Households", fill = "Disability (Head)"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

### E. By Age Group

```{r by-age}
ggplot(tenure_by_age_svy, aes(x = tenure_detailed, y = share, fill = age_head_group)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Survey-Weighted Tenure by Age Group of Household Head",
    x = "Tenure Type", y = "Share", fill = "Age Group"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

## V. Barriers to Homeownership

### A. Housing Cost Burden (Weighted)
`


## VI. First-Time Homebuyers and Programs

To understand how well state homebuyer assistance programs align with the needs of Californians, we assess the potential reach of several major programs, including:

- CalHFA First-Time Homebuyer Assistance (down payment/closing cost support),
- Forgivable Equity Builder Loan (for low-income first-time buyers), and
- Mortgage Credit Certificate (MCC) (tax credit for qualified buyers).

While administrative data on program recipients is limited and IPUMS microdata does not directly track homebuying history, we use proxy indicators to identify likely eligible households. In particular, we define a potential first-time buyer as:

- A renter,
- Between 25 and 44 years old, and
- Living in a low- to moderate-income household (under 120% of Area Median Income).

Our goal is to estimate how many households in California could qualify for these programs under current eligibility rules. We also disaggregate these estimates by race/ethnicity, income level, and region to evaluate who might be left out. 

### A.Define Eligibility Flags
```{r eligibility-flags}
data <- data %>%
  mutate(
    likely_ftb_eligible = ifelse(
      tenure_detailed == "Rented" &
      age_head >= 25 & age_head <= 44 &
      percent_ami <= 1.2,
      1, 0
    ),
    myhome_eligible = ifelse(likely_ftb_eligible == 1 & percent_ami <= 1.2, 1, 0),
    feb_eligible = ifelse(likely_ftb_eligible == 1 & percent_ami <= 0.8, 1, 0)
  )

if("OCC" %in% names(data)) {
  data <- data %>%
    mutate(
      farmworker_flag = ifelse(OCC %in% c(6010, 6050), 1, 0),
      joe_serna_eligible = ifelse(likely_ftb_eligible == 1 & farmworker_flag == 1, 1, 0)
    )
}
```

### B. Summary Table by Program and Subgroup
```{r table-by-subgroup}
# By income category
ftb_income <- svyby(~likely_ftb_eligible + myhome_eligible + feb_eligible,
                    ~income_category,
                    svy_households, svymean) %>%
  janitor::clean_names()

# By race/ethnicity
ftb_race <- svyby(~likely_ftb_eligible + myhome_eligible + feb_eligible,
                  ~race_ethnicity_head,
                  svy_households, svymean) %>%
  janitor::clean_names()

# Display tables
kable(ftb_income, digits = 3)
kable(ftb_race, digits = 3)
```

### C. Plots of Eligibility
```{r plot-ftb-by-income}
ggplot(ftb_income, aes(x = income_category, y = likely_ftb_eligible)) +
  geom_col(fill = "steelblue") +
  labs(title = "Likely First-Time Buyer Eligibility by Income Group",
       x = "Income Category", y = "Share Eligible") +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal()
```

```{r plot-feb-by-race}
ggplot(ftb_race, aes(x = race_ethnicity_head, y = feb_eligible, fill = race_ethnicity_head)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Forgivable Equity Builder Loan Eligibility by Race/Ethnicity",
       x = "Race/Ethnicity", y = "Share Eligible") +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal()
```

### D. Mapping Prep
```{r mapping-prep, eval=FALSE}
ftb_by_puma <- data %>%
  group_by(puma) %>%
  summarise(potential_ftbs = sum(weight_household * likely_ftb_eligible, na.rm = TRUE))

saveRDS(ftb_by_puma, here("data", "clean", "ftb_by_puma.rds"))
```


## VII. Summary Tables

### A. Average Income and Housing Costs

```{r summary_averages_new}
# Summary table of average income and housing costs
summary_stats <- households %>%
  summarise(
    avg_income = weighted.mean(income, weight_household, na.rm = TRUE),
    avg_owncost = weighted.mean(owncost, weight_household, na.rm = TRUE)
  )

knitr::kable(summary_stats, caption = "Weighted Averages of Household Income and Owner Costs")
```

```{r summary_averages_new}
# TODO: add summary table of income and owncost
```

## IV. Income Trends by Ownership Status

### A. Summary Table: Average Income by Tenure and multyear (Lara. RR)
```{r income-summary-table}
income_summary <- data %>%
  filter(multyear %in% c(2014, 2023),
         tenure_general %in% c("Owned", "Rented")) %>%
  mutate(income = as.numeric(as.character(income))) %>%
  group_by(multyear, tenure_general) %>%
  summarise(avg_income = mean(income, na.rm = TRUE),
            .groups = "drop")

print(income_summary)
```

### B. Line Plot: Average Income Over Time by Tenure (Lara. Updated)
```{r income-line-plot}
# ensure income is numeric
data$income <- as.numeric(as.character(data$income))

# filter for the correct tenure labels
data_filtered <- data %>%
  filter(tenure_general %in% c("Owner", "Renter")) %>%
  group_by(multyear, tenure_general) %>%
  summarise(
    mean_income   = mean(income, na.rm = TRUE),
    .groups       = "drop"
  )

#checking sample size weighted and unweighted
overall_counts_tenure <- data %>%
  # same filters as for the plot
  filter(tenure_general %in% c("Owner", "Renter")) %>%
  summarise(
    unweighted_n = n(),
    weighted_n   = sum(weight_household_norm, na.rm = TRUE),
    .groups      = "drop"
  )
print(overall_counts_tenure)

# weighted mean by year & tenure
data_filtered <- data %>%
  filter(tenure_general %in% c("Owner","Renter")) %>%
  group_by(multyear, tenure_general) %>%
  summarise(
    mean_income = weighted.mean(income, weight_household_norm, na.rm=TRUE), #(Karla) now that i updated to weight_household_norm im getting a weird # for weighted n
    .groups     = "drop"
  )

# plot
ggplot(data_filtered, aes(multyear, mean_income, colour = tenure_general)) +
  geom_rect(aes(xmin = 2018, xmax = 2023, ymin = -Inf, ymax = Inf),
            fill = "lightyellow", alpha = 0.1, color = NA) +  # Yellow region behind
  geom_vline(xintercept = 2020,
             linetype   = "dotted",
             colour     = "lightgray",
             size       = 1) +                                # pandemic marker
  geom_line(linewidth = 1) +
  
  ## axes
  scale_x_continuous(
    breaks = seq(2014, 2023, 2),
    limits = c(2014, 2023)) +
  scale_y_continuous(
    limits = c(30000, 130000),
    breaks = seq(30000, 130000, 20000),
    labels = function(x) x / 1000                       # show 40, 60, …
  ) +
  
  scale_color_manual(
    values = c("Owner" = "blue", "Renter" = "red"),
    name   = " "
  ) +
  labs(
    x     = "Year",
    y     = "Average Income (thousands of dollars)",
    title = "Avg. Income, Owners vs. Renters (2014 – 2023)"
  ) +
  theme_minimal() +
  theme(
    legend.position    = "right",
    plot.title         = element_text(hjust = 0.5)     # 0 = left, 0.5 = centre, 1 = right
  )
```
##(Lara. Updated)
```{r}
#Getting income stats for report analysis prose

# median income and percent change 
median_stats <- data %>%
  filter(tenure_general %in% c("Owner","Renter"),
         multyear %in% c(2014, 2023)) %>%
  group_by(tenure_general, multyear) %>%
  summarise(
    med = wtd.quantile(income, weight_household, probs = 0.5, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from   = multyear,
    names_prefix = "med",
    values_from  = med
  ) %>%
  mutate(
    pct = 100 * (med2023 - med2014) / med2014
  )

median_stats

#overall income and pct change
overall_pct <- data %>%
  filter(multyear %in% c(2014, 2023)) %>%
  group_by(multyear) %>%
  summarise(
    med = wtd.quantile(income, weight_household, probs = 0.5, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  summarise(
    pct_gain = 100 * (med[multyear == 2023] - med[multyear == 2014]) / med[multyear == 2014]
  ) %>%
  pull(pct_gain)

overall_pct
```
## V. Homeownership Trends by Sex


### A. Frequency and Summary Statistics (Lara. RR)
```{r freq-summary}
freq(data$gender_head)
freq(data$tenure_general)
freq(data$cost_burden_category)
summary(data$housing_cost_burden)
summary(data$housing_cost)
```

### B. Homeownership Summary by Sex (Lara. Updated)
```{r homeownership-rate-by-multyear-and-sex_head}
homeownership <- data %>%
  filter(
    gender_head  %in% c("Female", "Male"),
    tenure_general %in% c("Owner", "Renter")
  ) %>%
  group_by(multyear, gender_head) %>%
  summarise(
    prop_owned      = weighted.mean(tenure_general == "Owner",
                                    weight_household,
                                    na.rm = TRUE),
    prop_owned_pct  = prop_owned * 100,
    .groups         = "drop"
  )

head(homeownership)

#checking sample sizes
sex_sample_sizes <- data %>%
  filter(
    gender_head   %in% c("Female", "Male"),
    tenure_general %in% c("Owner", "Renter")
  ) %>%
  summarise(
    unweighted_n = n(),
    weighted_n   = sum(weight_household, na.rm = TRUE),
    .groups      = "drop"
  )

print(sex_sample_sizes)
```

### C. Trend Plot: Homeownership by Sex (Lara. Updated)

```{r}
# Plot the trends with the highlight and dotted vertical line.
ggplot(homeownership, aes(x = multyear, y = prop_owned_pct, color = gender_head, linetype = gender_head)) +
   # Highlight 2018-2024 period with a shaded region (behind everything)
 scale_y_continuous(limits = c(45, 65)) +
  geom_rect(aes(xmin = 2018, xmax = 2024, ymin = -Inf, ymax = Inf), 
            fill = "lightyellow", alpha = 0.1, color = NA) +  # Yellow region behind
  # Add a grey dotted vertical line at 2020.
  geom_vline(xintercept = 2020, linetype = "dotted", color = "lightgray", size = 1) +
  geom_line(linewidth = 1.2) +
  geom_point() +
  labs(x = "Year",
       y = "Homeownership Rate (%)",
       title = "Homeownership Trends by Sex of Head of Household (2014-2024)",
       color = "Sex of HH",
       linetype = "Sex of HH") +
  scale_color_manual(values = c("Female" = "purple", "Male" = "orange")) +
  scale_linetype_manual(values = c("Female" = "solid", "Male" = "solid")) +
  theme_minimal() +
  theme(legend.position = "right")
```

## VI. Housing Cost Burden Trends by Sex (Lara. Updated)

```{r homeownership-trends a}
#correcting housing cost variable by annualizing monthly cost
# coerce & drop IPUMS top-codes and compute annualized housing cost burden
data <- data %>%
  mutate(
    income       = as.numeric(income),
    housing_cost = as.numeric(housing_cost),
    income       = if_else(income >= 9999999, NA_real_, income),
    housing_cost = if_else(housing_cost >= 99999,   NA_real_, housing_cost),
    housing_cost_burden = if_else(
      income > 0,
      housing_cost * 12 / income,
      NA_real_
    )
  )

# sample size check for mortgage‐holders
overall_counts_gender <- data %>%
  filter(
    !is.na(housing_cost_burden),
    gender_head    %in% c("Female","Male"),
    tenure_detailed == "Owned - With Mortgage or Loan"
  ) %>%
  summarise(
    unweighted_n = n(),
    weighted_n   = sum(weight_household, na.rm = TRUE),
    .groups      = "drop"
  )
print(overall_counts_gender)

# compute median burden trends for mortgage‐holders 
data_gender_burden <- data %>%
  filter(
    tenure_detailed == "Owned - With Mortgage or Loan",
    !is.na(housing_cost_burden),
    gender_head %in% c("Female","Male")
  ) %>%
  mutate(group = paste(gender_head, "Mortgage Holder")) %>%
  group_by(multyear, group) %>%
  summarise(
    median_cost_burden = wtd.quantile(housing_cost_burden * 100,
                                      weight_household,
                                      probs = 0.5,
                                      na.rm = TRUE),
    .groups = "drop"
  )

# annotation x-position & compressed‐scale
x_label <- min(data_gender_burden$multyear) + diff(range(data_gender_burden$multyear)) / 4
compress_trans <- trans_new(
  "compress",
  transform = function(x) ifelse(x <= 50, x, 50 + (x - 50) / 10),
  inverse   = function(y) ifelse(y <= 50, y, (y - 50) * 10 + 50)
)

# plot using the weighted median
ggplot(data_gender_burden,
       aes(multyear, median_cost_burden, colour = group, linetype = group)) +
  geom_rect(aes(xmin = 2018, xmax = 2024, ymin = -Inf, ymax = Inf), 
            fill = "lightyellow", alpha = 0.1, color = NA) +  # Yellow region behind
  geom_vline(xintercept = 2020, linetype = "dotted", colour = "lightgray", size = 1) +
  geom_hline(yintercept = c(30, 50, 100), colour = c("green", "orange", "red")) +
  annotate("text", x = x_label, y = c(30, 50, 100),
           label  = c("Affordable (<30%)", "Moderate (30–50%)", "Severe (>50%)"),
           colour = c("green", "orange", "red"), vjust = -0.4) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_y_continuous(
    trans  = compress_trans,
    name   = "Median Housing Cost Burden (%)",
    limits = c(0, NA)
  ) +
  scale_colour_manual(values = c(
    "Female Mortgage Holder" = "purple",
    "Male Mortgage Holder"   = "orange"
  )) +
  scale_linetype_manual(values = c(
    "Female Mortgage Holder" = "solid",
    "Male Mortgage Holder"   = "solid"
  )) +
  labs(
    title    = "Median Housing Cost Burden for Mortgage Holders by Sex (2014–2023)",
    x        = "Year",
    colour   = "Group",
    linetype = "Group"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```
##(Lara. Updated)
```{r}
# compute average burden trends for mortgage holders
data_gender_burden_mean <- data %>%
  filter(
    tenure_detailed == "Owned - With Mortgage or Loan",
    !is.na(housing_cost_burden)
  ) %>%
  mutate(
    group       = paste(gender_head, "Mortgage Holder"),
    burden_pct  = housing_cost_burden * 100        # capped at 100%
  ) %>%
  group_by(multyear, group) %>%
  summarise(
    mean_cost_burden = weighted.mean(burden_pct, weight_household, na.rm = TRUE),
    .groups          = "drop"
  )

# position for annotations and compressed scale (same as before)
x_label <- min(data_gender_burden_mean$multyear) + diff(range(data_gender_burden_mean$multyear)) / 4
compress_trans <- trans_new(
  "compress",
  transform = function(x) ifelse(x <= 50, x, 50 + (x - 50) / 10),
  inverse   = function(y) ifelse(y <= 50, y, (y - 50) * 10 + 50)
)

# plot weighted‐mean burden
ggplot(data_gender_burden_mean,
       aes(multyear, mean_cost_burden, colour = group, linetype = group)) +
  annotate("rect", xmin = 2018, xmax = 2023, ymin = -Inf, ymax = Inf,
           fill = "lightyellow", alpha = 0.1) +
  geom_vline(xintercept = 2020, linetype = "dotted", colour = "lightgray", size = 1) +
  geom_hline(yintercept = c(30, 50, 100), colour = c("green", "orange", "red")) +
  annotate("text", x = x_label, y = c(30, 50, 100),
           label  = c("Affordable (<30%)","Moderate (30–50%)","Severe (>50%)"),
           colour = c("green","orange","red"), vjust = -0.4) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_y_continuous(
    trans  = compress_trans,
    name   = "Average Housing Cost Burden (%)",
    limits = c(0, NA)
  ) +
  scale_colour_manual(values = c(
    "Female Mortgage Holder" = "purple",
    "Male Mortgage Holder"   = "orange"
  )) +
  scale_linetype_manual(values = c(
    "Female Mortgage Holder" = "solid",
    "Male Mortgage Holder"   = "solid"
  )) +
  labs(
    title    = "Mean Housing Cost Burden for Mortgage Holders by Gender (2014–2023)",
    x        = "Year",
    colour   = "Group",
    linetype = "Group"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```
### A. Summary Table for Cost Burden (Owners) (Lara. Updated )
```{r}
data %>%
  filter(
    tenure_detailed == "Owned - With Mortgage or Loan",
    !is.na(housing_cost_burden)
  ) %>%
  group_by(multyear) %>%
  summarise(
    n             = n(),
    weighted_n    = sum(weight_household, na.rm=TRUE),
    mean_burden   = weighted.mean(housing_cost_burden, weight_household, na.rm=TRUE),
    median_burden = wtd.quantile(housing_cost_burden, weight_household, probs=0.5),
    p90_burden    = quantile(housing_cost_burden, 0.9, na.rm=TRUE),
    max_burden    = max(housing_cost_burden, na.rm=TRUE),
    .groups       = "drop"
  ) %>%
  print(n=Inf)
```

## XI. Ownership by Age and Race (MEG)

### A. Age Ownership Analysis (Meg. RR TO REVISE )
```{r age-cleaning-ownership}
# Convert 90+ to numeric and clean
data$age_head[data$age_head == "90 (90+ in 1980 and 1990)"] <- "90"

#make age_head numeric
data$age_head <- as.numeric(data$age_head)
# check
summary(data$age_head)

#drop NAs
data_clean <- data%>% filter(!is.na(age_head))
#note: dropped about 12,000 NA's here 

# Clean and recode tenure variable
data_clean <- data_clean %>%
  filter(multyear >= 2014 & multyear <= 2024,
         !is.na(tenure_general),
         tenure_general != "Other/Unknown") %>%
  mutate(tenure_general = recode(tenure_general, "Owned" = "Owner", "Rented" = "Renter"),
         # ownership dummy
         owner = if_else(tenure_general == "Owner", 1, 0))

# check table
prop.table(table(data_clean$owner))

```
### B. Ownership Rate by multyear (Meg. RR)
```{r}
# ownership rate by multyear
ownership_by_multyear <- data_clean %>%
  group_by(multyear) %>%
  summarise(ownership_rate = mean(owner, na.rm = TRUE))

ggplot(ownership_by_multyear, aes(x = multyear, y = ownership_rate)) +
  annotate("rect", xmin = 2018, xmax = 2023, ymin = -Inf, ymax = Inf, fill = "yellow", alpha = 0.1) +
geom_vline(xintercept = 2020, linetype = "dotted", color = "lightgrey", linewidth = 1) +
  geom_line(color = "#3b7EA1", size = 1.5) +
  geom_point(color = "#F4A261", size = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     expand = expansion(mult = c(1, 1))) +  # <- this adds 5% padding at top/bottom
  labs(x = "multyear", y = "Total Ownership Rate", title = "CA Homeownership Rate 2014-2023") +
  theme_minimal()

#detailed ownership rates by multyear
tenure_prop_table <- data_clean %>%
  group_by(multyear, tenure_detailed) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(multyear) %>%
  mutate(proportion = count / sum(count)) %>%
  select(multyear, tenure_detailed, proportion) %>%
  pivot_wider(names_from = multyear, values_from = proportion, values_fill = 0)
print(tenure_prop_table) 
#why is ownership so high

sum(data_clean$owner, na.rm=TRUE)

```

### C. Ownership Rate by Age and multyear (Meg. RR)
```{r}
#summarize ownership by age group and multyear
data_ownership <- data_clean %>%
  group_by(multyear, age_head_group)%>%
  summarise(
    ownership_rate = mean(owner, na.rm = TRUE),
    count = n(),
    .groups = "drop" )

# Fix group_by in weighted summary
data_ownership_weighted <- data_clean %>%
  filter(!is.na(owner), !is.na(weight_household), !is.na(age_head_group), age_head > 17) %>%
  group_by(multyear, age_head_group) %>%  # <- USE age_head_group
  summarise(
    ownership_rate = sum(owner * weight_household) / sum(weight_household),
    count = n(),
    .groups = "drop"
  )

# Fix plot aesthetics
plot_share_owner <- ggplot(data_ownership_weighted, aes(x = multyear, y = ownership_rate, color = age_head_group, group = age_head_group)) +
  annotate("rect", xmin = 2018, xmax = 2023, ymin = -Inf, ymax = Inf, fill = "yellow", alpha = 0.1) +
  geom_vline(xintercept = 2020, linetype = "dotted", color = "lightgrey", linewidth = 1) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_text(aes(label = scales::percent(ownership_rate, accuracy = 1)), 
            vjust = -0.5, size = 3, show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Homeownership Rates by Age Cohort (2014–2023)",
    x = "Year",
    y = "Ownership Rate",
    color = "Age Cohort"
  ) +
  scale_color_manual(values = c(  "#D9661f", "#3b7EA1", "#859438", "#EE1f60", "#9467bd")) +
  theme_minimal() +
  theme(legend.position = "bottom")


print(plot_share_owner)

```

### D. Change in Ownership Rate  (Meg. RR)

```{r}
#show change in ownership rate (since first multyear. should we use a different starting place?)
data_ownership_share <- data_ownership_weighted %>%
  group_by(age_head_group) %>%
  arrange(multyear) %>%
  mutate(change_from_start = ownership_rate - first(ownership_rate))

plot_change_in_owner <- ggplot(data_ownership_share, aes(x = multyear, y = change_from_start, color = age_head_group, group = age_head_group)) + annotate("rect", xmin = 2018, xmax = 2023, ymin = -Inf, ymax = Inf, fill = "yellow", alpha = 0.1) +
geom_vline(xintercept = 2020, linetype = "dotted", color = "lightgrey", linewidth = 1) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Change in Homeownership Rate by Age Cohort (vs. Starting multyear)",
    x = "multyear",
    y = "Change from Starting Ownership Rate",
    color = "Age Cohort"
  ) +
   scale_color_manual(values = c("#D9661f", "#3b7EA1", "#859438", "#EE1f60"))  +
  theme_minimal() +
  theme(legend.position = "bottom")

ggsave("change_in_ownership_rate.png", plot = plot_change_in_owner, width = 8, height = 6, dpi = 300)
```

### E. Average Age of Owners vs Renter (Weighted)  (Meg. RR.)
```{r}
# Check :confirm that renters have valid age and weight value
data_clean %>%
  filter(tenure_general == "Renter") %>%
  summarise(
    total_rows = n(),
    missing_age = sum(is.na(age_head)),
    missing_weight = sum(is.na(weight_household)),
    nonmissing_both = sum(!is.na(age_head) & !is.na(weight_household))
  )

renter_summary <- data_clean %>%
  filter(tenure_general == "Renter", !is.na(age_head), !is.na(weight_household)) %>%
  group_by(multyear) %>%
  summarise(
    avg_age = sum(age_head * weight_household) / sum(weight_household),
    count = n(),
    .groups = "drop"
  )

print(renter_summary)

```

```{r}
# Clean and ensure tenure_general is a proper character/factor
data_clean <- data_clean %>%
  mutate(tenure_general = as_factor(tenure_general))

# Summarize with weighted average
data_age_summary_weighted <- data_clean %>%
  filter(!is.na(age_head), !is.na(weight_household), !is.na(tenure_general)) %>%
  group_by(multyear, tenure_general) %>%
  summarise(
    avg_age = sum(age_head * weight_household) / sum(weight_household),
    count = n(),
    .groups = "drop"
  )


plot_age <- ggplot(data_age_summary_weighted, aes(x = multyear, y = avg_age, color = tenure_general, group = tenure_general)) +
  annotate("rect", xmin = 2018, xmax = 2023, ymin = -Inf, ymax = Inf, fill = "yellow", alpha = 0.1) +
  geom_vline(xintercept = 2020, linetype = "dotted", color = "lightgrey", linewidth = 1) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Average Age of Renters vs. Owners (2016–2023)",
    x = "Year",
    y = "Average Age of Head of Household",
    color = "Tenure Type"
  ) +
  scale_color_manual(values = c("Owner" = "#EE1f60", "Renter" = "#3b7EA1")) +
  scale_y_continuous(limits = c(30, 60)) +
  theme_minimal() +
  theme(legend.position = "bottom")
print(plot_age)

ggsave("avg_age_renters_vs_owners.png", plot = plot_age, width = 8, height = 6, dpi = 300)

```


## XII. Homeownership and Cost Burden by Race

### A. Weighted Ownership by Race (Meg)
```{r ownership-by-race}
data_summary_race_weighted <- data %>%
  filter(multyear >= 2014 & multyear <= 2023,
         !is.na(tenure_general),
         !is.na(race_ethnicity_head),
         !is.na(weight_household)) %>%
  mutate(owner = if_else(tenure_general == "Owned", 1, 0)) %>%
  group_by(multyear, race_ethnicity_head) %>%
  summarise(
    ownership_rate = sum(owner * weight_household) / sum(weight_household),
    count = n(),
    .groups = "drop"
  )

ggplot(data_summary_race_weighted, aes(x = multyear, y = ownership_rate, color = race_ethnicity_head)) +
  annotate("rect", xmin = 2018, xmax = 2023, ymin = -Inf, ymax = Inf, fill = "yellow", alpha = 0.1) +
  geom_vline(xintercept = 2020, linetype = "dotted", color = "lightgrey", linewidth = 1) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "CA Homeownership Rates by Race (2014–2023)",
       x = "multyear", y = "Ownership Rate", color = "Race") +
  theme_minimal() +
  theme(legend.position = "right")
```

## XIII Housing Burden (KARLA)

#### 1. Trend Over Time by AMI Category
Shows how cost burden changed over time within each AMI group.
```{r plot-burden-by-ami}
households %>%
  filter(multyear >= 2018, !is.na(ami_category), !is.na(housing_cost_burden)) %>%
  mutate(burden_flag = housing_cost_burden > 0.3) %>%
  group_by(multyear, ami_category) %>%
  summarise(overburdened = weighted.mean(burden_flag, weight_household, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = multyear, y = overburdened, color = ami_category)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Cost Burden Over Time by AMI Category (2018–2023)",
    x = "Year", y = "Percent Overburdened", color = "AMI Category"
  ) +
  theme_minimal()

```
#### 2. Boxplot of Cost Burden Distribution
Shows how much variation there is in cost burden within each AMI level.
```{r}
households %>%
  filter(!is.na(ami_category), !is.na(housing_cost_burden)) %>%
  ggplot(aes(x = ami_category, y = housing_cost_burden)) +
  geom_boxplot(fill = "#3b7ea1", alpha = 0.7) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Distribution of Housing Cost Burden by AMI Category",
    x = "AMI Category", y = "Housing Cost Burden (as % of Income)"
  ) +
  theme_minimal()

```
#### 3. Population Share Annotations
Still want to show the average overburden rate by AMI, but now with context on group size.
```{r}
households %>%
  filter(!is.na(ami_category), !is.na(housing_cost_burden)) %>%
  mutate(burden_flag = housing_cost_burden > 0.3) %>%
  group_by(ami_category) %>%
  summarise(
    n = sum(weight_household, na.rm = TRUE),
    overburdened = weighted.mean(burden_flag, weight_household, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("n = ", format(round(n), big.mark = ","))) %>%
  ggplot(aes(x = reorder(ami_category, -overburdened), y = overburdened)) +
  geom_col(fill = "#F4A261") +
  geom_text(aes(label = label), hjust = -0.1, size = 3.2) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Overburden Rate by AMI Category (with Population Size)",
    x = "AMI Category", y = "Percent Overburdened"
  ) +
  theme_minimal()
```
#### 4. Burden by AMI Category (selected years & bar)
```{r ami-cost-burden-stacked}
plot_data <- households %>%
  filter(multyear %in% c(2018, 2021, 2023),
         !is.na(ami_category),
         !is.na(cost_burden_category)) %>%
  group_by(multyear, ami_category, cost_burden_category) %>%
  summarise(households = sum(weight_household, na.rm = TRUE), .groups = "drop") %>%
  group_by(multyear, ami_category) %>%
  mutate(share = households / sum(households)) %>%
  ungroup()

ggplot(plot_data, aes(x = ami_category, y = share, fill = cost_burden_category)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  facet_wrap(~ multyear, nrow = 1) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(
    values = c("Affordable (<30%)" = "#F8766D",
               "Moderate (30-50%)" = "#00BFC4",
               "Severe (>50%)" = "#C77CFF",
               "Missing" = "gray70"),
    drop = FALSE
  ) +
  labs(
    title = "Cost Burden Distribution by AMI Category (2018, 2021, 2023)",
    x = "AMI Category", y = "Share of Households", fill = "Burden Level"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )

```
####  5.severe burden rate over time
```{r}
severe_burden <- households %>%
  filter(cost_burden_category == "Severe (>50%)",
         !is.na(ami_category)) %>%
  group_by(multyear, ami_category) %>%
  summarise(share_severe = sum(weight_household, na.rm = TRUE) / 
              sum(weight_household[!is.na(cost_burden_category)]), .groups = "drop")

ggplot(severe_burden, aes(x = multyear, y = share_severe, color = ami_category)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Severe Cost Burden Rates Over Time by AMI Category",
    x = "Year", y = "Percent Severely Burdened", color = "AMI Category"
  ) +
  theme_minimal()

```

#### 5. Burden by Race 
````{r}
cost_burden_summary <- households %>%
  filter(!is.na(housing_cost_burden),
         !is.na(weight_household),
         !is.na(race_ethnicity_head),
         housing_cost_burden <= 1) %>%  # remove bad values
  mutate(burden_flag = housing_cost_burden > 0.3) %>%
  group_by(race_ethnicity_head) %>%
  summarise(
    overburdened = weighted.mean(burden_flag, weight_household, na.rm = TRUE),
    count = n(),
    .groups = "drop"
  )

ggplot(cost_burden_summary, aes(x = reorder(race_ethnicity_head, -overburdened), y = overburdened)) +
  geom_col(fill = "#3b7EA1") +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  labs(
    title = "Share of Households Overburdened by Housing Costs",
    y = "Percent Overburdened", x = "Race/Ethnicity"
  ) +
  theme_minimal()


````

#### 6. Cost Burden by Race and Year (2018–2023)
```{r cost-burden-by-race-time}
# Step 1: Ensure correct data types
households <- households %>%
  mutate(
    housing_cost_burden = as.numeric(housing_cost_burden),
    weight_household = as.numeric(weight_household)
  )

# Step 2: Compute overburdened flag and summarize by year and race
cost_burden_by_multyear_race <- households %>%
  filter(multyear >= 2018,
         !is.na(housing_cost_burden),
         !is.na(race_ethnicity_head),
         !is.na(weight_household)) %>%
  mutate(burden_flag = housing_cost_burden > 0.3) %>%
  group_by(multyear, race_ethnicity_head) %>%
  summarise(overburdened = weighted.mean(burden_flag, weight_household, na.rm = TRUE),
            .groups = "drop")

# Step 3: Plot with optional annotation for 2020
ggplot(cost_burden_by_multyear_race,
       aes(x = multyear, y = overburdened, color = race_ethnicity_head)) +
  annotate("rect", xmin = 2020, xmax = 2021, ymin = 0, ymax = 1,
           fill = "lightgray", alpha = 0.1) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "Cost Burden by Race and Year (2018–2023)",
       x = "Year",
       y = "Percent Overburdened",
       color = "Race/Ethnicity") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")


```

## XIV. Time Series: Ownership and Cost Burden Trends
```{r homeownership-trends}
home_by_multyear <- households %>%
  filter(!is.na(tenure_general), multyear >= 2018) %>%
  mutate(homeowner = tenure_general == "Owned") %>%
  group_by(multyear) %>%
  summarise(home_own_rate = weighted.mean(homeowner, weight_household))

ggplot(home_by_multyear, aes(x = multyear, y = home_own_rate)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Homeownership Rate in California (2018–2023)",
       y = "Homeownership Rate", x = "Year")
```



