---
title: 'IPA'
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

## I. Project Overview

This file prepares household-level ACS data from IPUMS for analysis of
housing cost burden and ownership trends in California.

**Key assumptions:**

-   Only the householder (PERNUM == 1) is used to represent
    household-level traits
-   Household and person weights are already scaled and applied as
    provided
-   ACS uses complex sampling; analysis accounts for clustering using
    `survey::svydesign`
-   Data for 2020 may have quality issues due to COVID-19 disruptions
-   Income and housing costs are already CPI-adjusted in
    \_CPIU_2010variables - Using : Race: Simplified race/ethnicity
    identification [of Location of household head]

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## II. Set Up 

### A. Load Libraries

```{r warning =FALSE, message =FALSE}

library(survey)      
library(scales)      
library(ggthemes)    
library(ipumsr)
library(tidyverse)
library(here)
library(pastecs)
library(furniture)
library(gridExtra)
library(readxl)
library(writexl)
library(lubridate)
library(modeest)
library(stringr)
library(fastLink)
library(stargazer)
library(psych)
library(tibble)
library(infer)
library(purrr)
library(ggrepel)
library(latex2exp)
library(haven)
library(data.table)
```

### B. Clear memory

```{r}
rm(list = ls(all.names = TRUE))
gc()
```

### C. Custom Functions

```{r}
#some self-created functions that I frequently use
freq <- function(x) {
frequency <- table(x, useNA = "ifany")
proportion <- round(prop.table(table(x, useNA = "ifany"),),3)
cbind(frequency, proportion)
}

freq_sort <- function(x) {
frequency <- table(x, useNA = "ifany")
proportion <- round(prop.table(table(x, useNA = "ifany"),),3)
out <- cbind(frequency, proportion)
out[order(out[,1],decreasing=TRUE),]
}

freq2 = function(x,y){
table(x,y, useNA = "ifany")
}

freq2_col <-function(x,y){
prop.table(table(x,y,useNA = "ifany"),2)
}

freq2_row <-function(x,y){
prop.table(table(x,y,useNA = "ifany"),1)
}

  ttable <- function(dependent, by_variable){
  level <- c("By variable=0", "By variable=1", "Difference")
  means <- round(c(mean(dependent[by_variable==0], na.rm=TRUE),
  mean(dependent[by_variable==1], na.rm=TRUE),
  mean(dependent[by_variable==1], na.rm=TRUE) -
  mean(dependent[by_variable==0], na.rm=TRUE)),3)
  se_0 <- sd(dependent[by_variable==0],
  na.rm=TRUE)/sqrt(sum(!is.na(dependent[by_variable==0])))
  se_1 <- sd(dependent[by_variable==1],
  na.rm=TRUE)/sqrt(sum(!is.na(dependent[by_variable==1])))
  se_diff <-sqrt(se_0^2+ se_1^2)
  se <- round(c(se_0, se_1, se_diff),3)
  df_correction <- c(1,1,2)
  N <- c(sum(!is.na(dependent[by_variable==0])) ,
  sum(!is.na(dependent[by_variable==01])), sum(!is.na(dependent)))
  results <- as.data.frame(cbind(level, means, se, N, df_correction))
  results$means <-as.numeric(results$means)
  results$se <-as.numeric(results$se)
  results$N <-as.numeric(results$N)
  results$df_correction <-as.numeric(results$df_correction)
  ttable <- results %>%
  mutate(t_stat = round(means/se,3),
  df_t = N - df_correction,
  p_value = 2*pt(abs(t_stat), df=df_t, lower.tail=FALSE)
  ) %>%
  select(level, means, se, t_stat, df_t, p_value )
  rm(level, means, se_0, se_1, se_diff, se, df_correction, N, results)
  ttable
}

```

## III. Load IPUMS data 

### A. Import Data and DDI File
```{r}
mem.maxVSize()
if (!require("ipumsr")) stop("Reading IPUMS data into R requires the ipumsr package. It can be installed using the following command: install.packages('ipumsr')")
ddi <- read_ipums_ddi("usa_00003.xml")
data <- read_ipums_micro(ddi)
print(object.size(data), units = "GB")
```

### B. Rename Variables to Lowercase
```{r}
# rename vars to lower
names(data) <- tolower(names(data))
```

## IV. Descriptive Checks

### A. General Summary
```{r}
# summary stats for the whole dataset 
summary(data)
```


### B. Frequency Checks
Notes:

-   ownershp: Ownership of dwelling (tenure) [general version]

-   ownershpd:Ownership of dwelling (tenure) [detailed version]

-   Year= Census Year

-   For the multi-year ACS/PRCS samples, YEAR indicates the last year of
    data included

-   For the actual year of survey in these multi-year data, see
    MULTYEAR.

-   MultYear = IPUMS year of survey , multi-year ACS/PRCS

    -   MULTYEAR identifies the actual year of survey in multi-year
        ACS/PRCS samples.
    -   MULTYEAR gives the single-year sample from which the case was
        drawn (2005, 2006, or 2007 for the 2005-2007 3-year data; 2006,
        2007, or 2008 for the 2006-2008 3-year data; and so on).

-   The distribution of records per year varies in the multiyear var
    because we are looking at rolling samples.So for exmaple, the year
    2015 will have records from the 2014-2018 and 2015-2019 sample. Thus
    2014 and 2023 only have 1 year of data.


```{r freq-checks}
# age
freq(data$age)

#state
freq(data$statefip)

# race
freq(data$race)

freq(data$race_head)

# owenership
freq(data$ownershp)

freq(data$ownershpd)

# dissability
freq(data$diffcare)

freq(data$diffphys)

freq(data$diffrem)

freq(data$diffmob)

freq(data$diffeye)

freq(data$diffhear)

# income
summary(data$hhincome)
summary(data$hhincome_cpiu_2010)


#yeare
print(data %>% count(year))
print(data %>% count(multyear))
```



## V. Data Cleaning and Variable Setup

### A. Create Identifiers and Key Variables

Notes:

-   SERIAL: is an identifying number unique to each household record in
    a given sample. All person records are assigned the same serial
    number as the household record that they follow.

-   PERNUM numbers all persons within each household consecutively in
    the order in which they appear on the original census or survey
    form. When combined with SAMPLE and SERIAL, PERNUM uniquely
    identifies each person within the IPUMS.

-   HHWT is a 6-digit numeric variable which indicates how many
    households in the U.S. population are represented by a given
    household in an IPUMS sample and has two implied decimals. For
    example, a HHWT value of 010461 should be interpreted as 104.61.
    HHWT specific variable codes for missing, edited, or unidentified
    observations, observations not applicable (N/A), observations not in
    universe (NIU), top and bottom value coding, etc. are provided below
    if applicable by Census year (and data sample if specified).
    However, our HHWT values are showing as numeric values like 26,
    34, 38. This ️likely means that the weights were imported already
    scaled correctly (i.e., the implied decimals have already been
    applied). So we will not weigh_person/household as recommended above
    hhwt/100.

-   Generate person identifier: household identifier + \# that
    identifies the person within the household

```{r}
data_clean <- data %>%
  mutate(
    household_id = serial, # 
    person_id = paste(serial, pernum, sep = "-"), 
    income = hhincome_cpiu_2010,
    housing_cost = owncost_cpiu_2010, 
    weight_person = perwt,
    weight_household = hhwt
  )

# remove original data(can read after, I just don't have memory)
rm(data)
```

### B. Set Up Survey Designs

-   The ACS data isn't collected randomly — it uses a complex design
    where some types of households are more likely to be selected than
    others, and households are grouped (clustered) by geography.

-   If we ignore this, our results might look more precise than they
    really are (i.e., too small standard errors).

-   Using svydesign() helps R understand how the data was collected so
    it can:

    -   Properly apply the weights (so each household counts for the
        right number of people),
    -   Adjust for clustering (so confidence intervals and p-values are
        more realistic).

-   We want to use this design for:

    -   Calculating means or proportions with standard errors
    -   Plotting trends with confidence intervals
    -   Running regressions or hypothesis tests

-   CLUSTER: Household cluster for variance estimation

**Household level survey design**
```{r}
hh_design <- svydesign(
  id = ~cluster,
  weights = ~weight_household,
  data = data_clean
)
```

**person level survey design**
```{r}
person_design <- svydesign(
  id = ~cluster,
  weights = ~perwt,
  data = data_clean
)
```


### C. Clean and Recode Variables

- Age
- Race & Ethnicity
- Veteran Status
- Disability
- Tenure (Ownership)
- Household Income Brackets
- Housing Cost Burden
- Household Size

#### 1. Age

Person
```{r}
# clean age 
data_clean <- data_clean %>%
  mutate(
    age_group = cut(age, breaks = c(0, 17, 34, 54, 74, Inf), 
                    labels = c("0-17", "18-34", "35-54", "55-74", "75+"))
  )
freq(data_clean$age_group)
```

House
```{r}
data_clean <- data_clean %>%
  mutate(
    age_head_group = cut(age, breaks = c(0, 17, 34, 54, 74, Inf), 
                         labels = c("0-17", "18-34", "35-54", "55-74", "75+"))
  )

```

#### 2. Race & ethnicity

Notes: 

- Recode Hispanic origin and race/ethnicity using numeric codes
- HISPAN: 0 = Not Hispanic

Person
```{r}
data_clean <- data_clean %>%
  mutate(
    race_group = case_when(
      race == 1 ~ "White",
      race == 2 ~ "Black",
      race == 3 ~ "Native American",
      race %in% 4:6 ~ "Asian/Pacific Islander",
      race %in% 7:8 ~ "Multiracial",
      race == 9 ~ "Other",
      TRUE ~ "Unknown"
    )
  )


data_clean <- data_clean %>%
  mutate(
    race_ethnicity = case_when(
      hispan != 0 ~ "Hispanic",
      hispan == 0 & race_group == "White" ~ "Non-Hispanic White",
      hispan == 0 & race_group == "Black" ~ "Non-Hispanic Black",
      hispan == 0 & race_group == "Asian/Pacific Islander" ~ "Non-Hispanic Asian/Pacific Islander",
      hispan == 0 & race_group == "Native American" ~ "Non-Hispanic Native American",
      hispan == 0 & race_group == "Multiracial" ~ "Non-Hispanic Multiracial",
      hispan == 0 & race_group == "Other" ~ "Non-Hispanic Other",
      TRUE ~ "Unknown"
    )
  )

```

Household
```{r}
data_clean <- data_clean %>%
  mutate(
    race_group_head = case_when(
      race_head == 1 ~ "White",
      race_head == 2 ~ "Black",
      race_head == 3 ~ "Native American",
      race_head %in% 4:6 ~ "Asian/Pacific Islander",
      race_head %in% 7:8 ~ "Multiracial",
      race_head == 9 ~ "Other",
      TRUE ~ "Unknown"
    ),
    
    race_ethnicity_head = case_when(
      hispan_head != 0 ~ "Hispanic",  # 0 = Not Hispanic
      hispan_head == 0 & race_group_head == "White" ~ "Non-Hispanic White",
      hispan_head == 0 & race_group_head == "Black" ~ "Non-Hispanic Black",
      hispan_head == 0 & race_group_head == "Asian/Pacific Islander" ~ "Non-Hispanic Asian/Pacific Islander",
      hispan_head == 0 & race_group_head == "Native American" ~ "Non-Hispanic Native American",
      hispan_head == 0 & race_group_head == "Multiracial" ~ "Non-Hispanic Multiracial",
      hispan_head == 0 & race_group_head == "Other" ~ "Non-Hispanic Other",
      TRUE ~ "Unknown"
    )
  )

freq(data_clean$race_ethnicity_head)
```

#### 3.Veteran Status

Person
```{r}
data_clean <- data_clean %>%
  mutate(
    veteran_person = case_when(
      vetstatd == 2 ~ "Veteran",
      vetstatd == 3 ~ "Active duty",
      vetstatd == 4 ~ "Reserve/National Guard",
      vetstatd == 1 ~ "Non-veteran",
      vetstatd == 0 ~ NA_character_,  # NIU (not in universe)
      TRUE ~ "Unknown"
    )
  )
freq(data_clean$veteran_person)
```

Household
```{r}
data_clean <- data_clean %>%
  mutate(
    veteran_head = case_when(
      vetstat_head == 1 ~ "Veteran",
      vetstat_head == 2 ~ "Non-veteran",
      vetstat_head == 0 ~ NA_character_,  # NIU (Not in Universe)
      TRUE ~ "Unknown"
    )
  )

freq(data_clean$veteran_head)
```

#### 4. Disability

Person
```{r}
data_clean <- data_clean %>%
  mutate(
    # 1 = Yes (has difficulty), 2 = No, 0 = NIU
    diffrem_flag = case_when(
      diffrem == 1 ~ 1,
      diffrem == 2 ~ 0,
      diffrem == 0 ~ NA_real_,  # Not in universe
      TRUE ~ NA_real_
    ),
    
    diffphys_flag = case_when(
      diffphys == 1 ~ 1,
      diffphys == 2 ~ 0,
      diffphys == 0 ~ NA_real_,
      TRUE ~ NA_real_
    ),
    
    diffmob_flag = case_when(
      diffmob == 1 ~ 1,
      diffmob == 2 ~ 0,
      diffmob == 0 ~ NA_real_,
      TRUE ~ NA_real_
    ),
    
    diffcare_flag = case_when(
      diffcare == 1 ~ 1,
      diffcare == 2 ~ 0,
      diffcare == 0 ~ NA_real_,
      TRUE ~ NA_real_
    ),
    
    diffsens_flag = case_when(
      diffsens == 1 ~ 1,
      diffsens == 2 ~ 0,
      diffsens == 0 ~ NA_real_,
      TRUE ~ NA_real_
    )
  )

data_clean <- data_clean %>%
  mutate(
    disability_person = case_when(
      rowSums(select(., diffrem_flag, diffphys_flag, diffmob_flag, diffcare_flag, diffsens_flag), na.rm = TRUE) >= 1 ~ "Yes",
      rowSums(select(., diffrem_flag, diffphys_flag, diffmob_flag, diffcare_flag, diffsens_flag), na.rm = TRUE) == 0 ~ "No",
      TRUE ~ NA_character_
    )
  )

freq(data_clean$disability_person)
```

Household
```{r}
# Disability flags for household head using numeric values
data_clean <- data_clean %>%
  mutate(
    diffrem_head_flag = case_when(
      diffrem_head == 1 ~ 1,
      diffrem_head == 2 ~ 0,
      diffrem_head == 0 ~ NA_real_,
      TRUE ~ NA_real_
    ),
    diffphys_head_flag = case_when(
      diffphys_head == 1 ~ 1,
      diffphys_head == 2 ~ 0,
      diffphys_head == 0 ~ NA_real_,
      TRUE ~ NA_real_
    ),
    diffmob_head_flag = case_when(
      diffmob_head == 1 ~ 1,
      diffmob_head == 2 ~ 0,
      diffmob_head == 0 ~ NA_real_,
      TRUE ~ NA_real_
    ),
    diffcare_head_flag = case_when(
      diffcare_head == 1 ~ 1,
      diffcare_head == 2 ~ 0,
      diffcare_head == 0 ~ NA_real_,
      TRUE ~ NA_real_
    ),
    diffsens_head_flag = case_when(
      diffsens_head == 1 ~ 1,
      diffsens_head == 2 ~ 0,
      diffsens_head == 0 ~ NA_real_,
      TRUE ~ NA_real_
    )
  )

data_clean <- data_clean %>%
  mutate(
    disability_head = case_when(
      rowSums(select(., diffrem_head_flag, diffphys_head_flag, diffmob_head_flag, diffcare_head_flag, diffsens_head_flag), na.rm = TRUE) >= 1 ~ "Yes",
      rowSums(select(., diffrem_head_flag, diffphys_head_flag, diffmob_head_flag, diffcare_head_flag, diffsens_head_flag), na.rm = TRUE) == 0 ~ "No",
      TRUE ~ NA_character_
    )
  )


freq(data_clean$disability_head)
```

#### 5. HH Ownership

Notes:

- OWNERSHP: indicates whether the housing unit was rented or owned by its inhabitants. Housing units acquired with a mortgage or other lending arrangement(s) are classified as "owned," even if repayment was not yet completed.

Detailed
```{r}
data_clean <- data_clean %>%
  mutate(
    tenure_detailed = case_when(
      ownershpd == 12 ~ "Owned - Free and Clear",
      ownershpd == 13 ~ "Owned - With Mortgage",
      ownershpd == 22 ~ "Rented - Cash Rent",
      ownershpd == 21 ~ "Rented - No Cash Rent",
      TRUE ~ "Other/Unknown"
    )
  )

count(data_clean, tenure_detailed)
```

General
```{r}
data_clean <- data_clean %>%
  mutate(
    tenure_general = case_when(
      str_detect(tenure_detailed, "Owned") ~ "Owned",
      str_detect(tenure_detailed, "Rented") ~ "Rented",
      TRUE ~ "Other/Unknown"
    )
  )

count(data_clean, tenure_general)

```

#### 6. Household income

```{r}
# create income brackets 
data_clean <- data_clean %>%
  mutate(
    income_bracket = case_when(
    income < 25000 ~ "< $25k",
    income >= 25000 & income < 50000 ~ "$25k - $49k",
    income >= 50000 & income < 75000 ~ "$50k - $74k",
    income >= 75000 & income < 100000 ~ "$75k - $99k",
    income >= 100000 ~ "$100k+",
    TRUE ~ "Unknown"
   )
  )
freq(data_clean$income_bracket)
```

#### 7. Housing burden

Notes:

-   Create a housing cost burden ratio: housing_cost_burden
    =monthly_housing_cost/ annual_household_income
-   If income is 0 or missing, it assigns NA to avoid divding by 0.
-   Selected monthly owner costs [standardized using CPIU_2010]

```{r}
data_clean <- data_clean %>%
  mutate(
    income = as.numeric(hhincome_cpiu_2010),
    housing_cost = as.numeric(owncost_cpiu_2010),
    housing_cost_burden = ifelse(income > 0, housing_cost / income, NA),
    cost_burden_category = case_when(
      is.na(housing_cost_burden) ~ "Missing",
      housing_cost_burden < 0.3 ~ "Affordable (<30%)",
      housing_cost_burden < 0.5 ~ "Moderate (30-50%)",
      housing_cost_burden >= 0.5 ~ "Severe (>50%)",
      TRUE ~ "Unknown"
    )
  )

freq(data_clean$cost_burden_category)
```

#### 8. Household size
```{r}
data_clean <- data_clean %>%
  mutate(
    household_size_group = case_when(
      numprec == 1 ~ "1 person",
      numprec == 2 ~ "2 people",
      numprec >= 3 & numprec <= 4 ~ "3-4 people",
      numprec >= 5 ~ "5+ people",
      TRUE ~ "Unknown"
  )
  )
freq(data_clean$household_size_group)
```

## VI. Export Cleaned Files
```{r}
# use readr - faster than write.csv
saveRDS(data_clean, "full_clean.rds")
```

## VII. Create Household-Level Subset

Notes:

- Aggregate household-level data
- Keep only household heads
- # as recommended by IPUMS: users should also be sure to select one person (e.g., PERNUM = 1) to represent the entire household.
- q* variables are data quality flags that indicate whether a value was allocated (imputed) by the Census Bureau
```{r}
households <- data_clean %>%
  filter(pernum == 1) %>%  
  select(
    household_id = serial,
    pernum,
    weight_household = hhwt,
    year, 
    multyear,
    race_ethnicity_head,
    veteran_head,
    disability_head,
    sex_head,
    age_head,
    age_head_group,
    income = hhincome_cpiu_2010,
    income_bracket,
    housing_cost = owncost_cpiu_2010,
    housing_cost_burden,
    cost_burden_category,
    tenure_detailed,
    tenure_general,
    met_area = met2013, 
    puma, 
    cluster,
    strata,
    gq,
    qhhincome, 
    qowncost, 
    qownersh,
  )%>% 
  mutate(
    any_imputed = qownersh == 1 | qhhincome == 1 | qowncost == 1 
  )

saveRDS(households, "household_clean.rds")

summary(households)
```

### A. HH level Checks

```{r}
households %>% count(race_ethnicity_head, tenure_detailed, wt = weight_household)
households %>% count(veteran_head, cost_burden_category, wt = weight_household)

```
