---
title: "IPA"
author: ''
date: "`r Sys.Date()`"
output:
  html_document:
    data_print: paged
  pdata_document: default
  word_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## I. Set Up 

### A. Load Libraries

```{r warning =FALSE, message =FALSE}

library(survey)      
library(scales)      
library(ggthemes)    
library(ipumsr)
library(tidyverse)
library(here)
library(pastecs)
library(furniture)
library(gridExtra)
library(readxl)
library(writexl)
library(lubridate)
library(modeest)
library(stringr)
library(fastLink)
library(stargazer)
library(psych)
library(tibble)
library(infer)
library(purrr)
library(ggrepel)
library(latex2exp)
library(haven)
library(data.table)
library(Hmisc)
library(matrixStats)
library(knitr)
library(kableExtra)


```

### B. Clear Memory
```{r}
#rm(list = ls(all.names = TRUE))
#gc()
```

### C. Custom Functions

```{r}
#some self-created functions that I frequently use
freq <- function(x) {
frequency <- table(x, useNA = "ifany")
proportion <- round(prop.table(table(x, useNA = "ifany"),),3)
cbind(frequency, proportion)
}

freq_sort <- function(x) {
frequency <- table(x, useNA = "ifany")
proportion <- round(prop.table(table(x, useNA = "ifany"),),3)
out <- cbind(frequency, proportion)
out[order(out[,1],decreasing=TRUE),]
}

freq2 = function(x,y){
table(x,y, useNA = "ifany")
}

freq2_col <-function(x,y){
prop.table(table(x,y,useNA = "ifany"),2)
}

freq2_row <-function(x,y){
prop.table(table(x,y,useNA = "ifany"),1)
}

  ttable <- function(dependent, by_variable){
  level <- c("By variable=0", "By variable=1", "Difference")
  means <- round(c(mean(dependent[by_variable==0], na.rm=TRUE),
  mean(dependent[by_variable==1], na.rm=TRUE),
  mean(dependent[by_variable==1], na.rm=TRUE) -
  mean(dependent[by_variable==0], na.rm=TRUE)),3)
  se_0 <- sd(dependent[by_variable==0],
  na.rm=TRUE)/sqrt(sum(!is.na(dependent[by_variable==0])))
  se_1 <- sd(dependent[by_variable==1],
  na.rm=TRUE)/sqrt(sum(!is.na(dependent[by_variable==1])))
  se_diff <-sqrt(se_0^2+ se_1^2)
  se <- round(c(se_0, se_1, se_diff),3)
  data_correction <- c(1,1,2)
  N <- c(sum(!is.na(dependent[by_variable==0])) ,
  sum(!is.na(dependent[by_variable==01])), sum(!is.na(dependent)))
  results <- as.data.frame(cbind(level, means, se, N, data_correction))
  results$means <-as.numeric(results$means)
  results$se <-as.numeric(results$se)
  results$N <-as.numeric(results$N)
  results$data_correction <-as.numeric(results$data_correction)
  ttable <- results %>%
  mutate(t_stat = round(means/se,3),
  data_t = N - data_correction,
  p_value = 2*pt(abs(t_stat), data=data_t, lower.tail=FALSE)
  ) %>%
  select(level, means, se, t_stat, data_t, p_value )
  rm(level, means, se_0, se_1, se_diff, se, data_correction, N, results)
  ttable
}

```

## II. Data Loading

### A. Load IPUMS data 
```{r}
#setwd("...")
data <-readRDS("~/Documents/IPA Proj/01_IPA_cleaning_household_clean.rds")

#print(object.size(data), units = "GB")
```

### B. Cast values as labels
```{r}
# Convert all labelled variables to factors with labels
data <- data %>%
  mutate(across(where(is.labelled), as_factor))
```

## III. Descriptive Overview

```{r }
#str(data)

names(data)

#summary(data)
```

## IV. Income Trends by Ownership Status

### A. Summary Table: Average Income by Tenure and multyear (Lara. RR)
```{r income-summary-table}
income_summary <- data %>%
  filter(multyear %in% c(2014, 2023),
         tenure_general %in% c("Owned", "Rented")) %>%
  mutate(income = as.numeric(as.character(income))) %>%
  group_by(multyear, tenure_general) %>%
  summarise(avg_income = mean(income, na.rm = TRUE),
            .groups = "drop")

print(income_summary)
```

### B. Line Plot: Average Income Over Time by Tenure (Lara. Updated)
```{r income-line-plot}
# ensure income is numeric
data$income <- as.numeric(as.character(data$income))

# filter for the correct tenure labels
data_filtered <- data %>%
  filter(tenure_general %in% c("Owner", "Renter")) %>%
  group_by(multyear, tenure_general) %>%
  summarise(
    mean_income   = mean(income, na.rm = TRUE),
    .groups       = "drop"
  )

#checking sample size weighted and unweighted
overall_counts_tenure <- data %>%
  # same filters as for the plot
  filter(tenure_general %in% c("Owner", "Renter")) %>%
  summarise(
    unweighted_n = n(),
    weighted_n   = sum(weight_household_norm, na.rm = TRUE),
    .groups      = "drop"
  )
print(overall_counts_tenure)

# weighted mean by year & tenure
data_filtered <- data %>%
  filter(tenure_general %in% c("Owner","Renter")) %>%
  group_by(multyear, tenure_general) %>%
  summarise(
    mean_income = weighted.mean(income, weight_household_norm, na.rm=TRUE), #(Karla) now that i updated to weight_household_norm im getting a weird # for weighted n
    .groups     = "drop"
  )

# plot
ggplot(data_filtered, aes(multyear, mean_income, colour = tenure_general)) +
  geom_rect(aes(xmin = 2018, xmax = 2023, ymin = -Inf, ymax = Inf),
            fill = "lightyellow", alpha = 0.1, color = NA) +  # Yellow region behind
  geom_vline(xintercept = 2020,
             linetype   = "dotted",
             colour     = "lightgray",
             size       = 1) +                                # pandemic marker
  geom_line(linewidth = 1) +
  
  ## axes
  scale_x_continuous(
    breaks = seq(2014, 2023, 2),
    limits = c(2014, 2023)) +
  scale_y_continuous(
    limits = c(30000, 130000),
    breaks = seq(30000, 130000, 20000),
    labels = function(x) x / 1000                       # show 40, 60, …
  ) +
  
  scale_color_manual(
    values = c("Owner" = "blue", "Renter" = "red"),
    name   = " "
  ) +
  labs(
    x     = "Year",
    y     = "Average Income (thousands of dollars)",
    title = "Avg. Income, Owners vs. Renters (2014 – 2023)"
  ) +
  theme_minimal() +
  theme(
    legend.position    = "right",
    plot.title         = element_text(hjust = 0.5)     # 0 = left, 0.5 = centre, 1 = right
  )
```
##(Lara. Updated)
```{r}
#Getting income stats for report analysis prose

# median income and percent change 
median_stats <- data %>%
  filter(tenure_general %in% c("Owner","Renter"),
         multyear %in% c(2014, 2023)) %>%
  group_by(tenure_general, multyear) %>%
  summarise(
    med = wtd.quantile(income, weight_household, probs = 0.5, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from   = multyear,
    names_prefix = "med",
    values_from  = med
  ) %>%
  mutate(
    pct = 100 * (med2023 - med2014) / med2014
  )

median_stats

#overall income and pct change
overall_pct <- data %>%
  filter(multyear %in% c(2014, 2023)) %>%
  group_by(multyear) %>%
  summarise(
    med = wtd.quantile(income, weight_household, probs = 0.5, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  summarise(
    pct_gain = 100 * (med[multyear == 2023] - med[multyear == 2014]) / med[multyear == 2014]
  ) %>%
  pull(pct_gain)

overall_pct
```
## V. Homeownership Trends by Sex


### A. Frequency and Summary Statistics (Lara. RR)
```{r freq-summary}
freq(data$gender_head)
freq(data$tenure_general)
freq(data$cost_burden_category)
summary(data$housing_cost_burden)
summary(data$housing_cost)
```

### B. Homeownership Summary by Sex (Lara. Updated)
```{r homeownership-rate-by-multyear-and-sex_head}
homeownership <- data %>%
  filter(
    gender_head  %in% c("Female", "Male"),
    tenure_general %in% c("Owner", "Renter")
  ) %>%
  group_by(multyear, gender_head) %>%
  summarise(
    prop_owned      = weighted.mean(tenure_general == "Owner",
                                    weight_household,
                                    na.rm = TRUE),
    prop_owned_pct  = prop_owned * 100,
    .groups         = "drop"
  )

head(homeownership)

#checking sample sizes
sex_sample_sizes <- data %>%
  filter(
    gender_head   %in% c("Female", "Male"),
    tenure_general %in% c("Owner", "Renter")
  ) %>%
  summarise(
    unweighted_n = n(),
    weighted_n   = sum(weight_household, na.rm = TRUE),
    .groups      = "drop"
  )

print(sex_sample_sizes)
```

### C. Trend Plot: Homeownership by Sex (Lara. Updated)

```{r}
# Plot the trends with the highlight and dotted vertical line.
ggplot(homeownership, aes(x = multyear, y = prop_owned_pct, color = gender_head, linetype = gender_head)) +
   # Highlight 2018-2024 period with a shaded region (behind everything)
 scale_y_continuous(limits = c(45, 65)) +
  geom_rect(aes(xmin = 2018, xmax = 2024, ymin = -Inf, ymax = Inf), 
            fill = "lightyellow", alpha = 0.1, color = NA) +  # Yellow region behind
  # Add a grey dotted vertical line at 2020.
  geom_vline(xintercept = 2020, linetype = "dotted", color = "lightgray", size = 1) +
  geom_line(linewidth = 1.2) +
  geom_point() +
  labs(x = "Year",
       y = "Homeownership Rate (%)",
       title = "Homeownership Trends by Sex of Head of Household (2014-2024)",
       color = "Sex of HH",
       linetype = "Sex of HH") +
  scale_color_manual(values = c("Female" = "purple", "Male" = "orange")) +
  scale_linetype_manual(values = c("Female" = "solid", "Male" = "solid")) +
  theme_minimal() +
  theme(legend.position = "right")
```

## VI. Housing Cost Burden Trends by Sex (Lara. Updated)

```{r homeownership-trends a}
#correcting housing cost variable by annualizing monthly cost
# coerce & drop IPUMS top-codes and compute annualized housing cost burden
data <- data %>%
  mutate(
    income       = as.numeric(income),
    housing_cost = as.numeric(housing_cost),
    income       = if_else(income >= 9999999, NA_real_, income),
    housing_cost = if_else(housing_cost >= 99999,   NA_real_, housing_cost),
    housing_cost_burden = if_else(
      income > 0,
      housing_cost * 12 / income,
      NA_real_
    )
  )

# sample size check for mortgage‐holders
overall_counts_gender <- data %>%
  filter(
    !is.na(housing_cost_burden),
    gender_head    %in% c("Female","Male"),
    tenure_detailed == "Owned - With Mortgage or Loan"
  ) %>%
  summarise(
    unweighted_n = n(),
    weighted_n   = sum(weight_household, na.rm = TRUE),
    .groups      = "drop"
  )
print(overall_counts_gender)

# compute median burden trends for mortgage‐holders 
data_gender_burden <- data %>%
  filter(
    tenure_detailed == "Owned - With Mortgage or Loan",
    !is.na(housing_cost_burden),
    gender_head %in% c("Female","Male")
  ) %>%
  mutate(group = paste(gender_head, "Mortgage Holder")) %>%
  group_by(multyear, group) %>%
  summarise(
    median_cost_burden = wtd.quantile(housing_cost_burden * 100,
                                      weight_household,
                                      probs = 0.5,
                                      na.rm = TRUE),
    .groups = "drop"
  )

# annotation x-position & compressed‐scale
x_label <- min(data_gender_burden$multyear) + diff(range(data_gender_burden$multyear)) / 4
compress_trans <- trans_new(
  "compress",
  transform = function(x) ifelse(x <= 50, x, 50 + (x - 50) / 10),
  inverse   = function(y) ifelse(y <= 50, y, (y - 50) * 10 + 50)
)

# plot using the weighted median
ggplot(data_gender_burden,
       aes(multyear, median_cost_burden, colour = group, linetype = group)) +
  geom_rect(aes(xmin = 2018, xmax = 2024, ymin = -Inf, ymax = Inf), 
            fill = "lightyellow", alpha = 0.1, color = NA) +  # Yellow region behind
  geom_vline(xintercept = 2020, linetype = "dotted", colour = "lightgray", size = 1) +
  geom_hline(yintercept = c(30, 50, 100), colour = c("green", "orange", "red")) +
  annotate("text", x = x_label, y = c(30, 50, 100),
           label  = c("Affordable (<30%)", "Moderate (30–50%)", "Severe (>50%)"),
           colour = c("green", "orange", "red"), vjust = -0.4) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_y_continuous(
    trans  = compress_trans,
    name   = "Median Housing Cost Burden (%)",
    limits = c(0, NA)
  ) +
  scale_colour_manual(values = c(
    "Female Mortgage Holder" = "purple",
    "Male Mortgage Holder"   = "orange"
  )) +
  scale_linetype_manual(values = c(
    "Female Mortgage Holder" = "solid",
    "Male Mortgage Holder"   = "solid"
  )) +
  labs(
    title    = "Median Housing Cost Burden for Mortgage Holders by Sex (2014–2023)",
    x        = "Year",
    colour   = "Group",
    linetype = "Group"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```
##(Lara. Updated)
```{r}
# compute average burden trends for mortgage holders
data_gender_burden_mean <- data %>%
  filter(
    tenure_detailed == "Owned - With Mortgage or Loan",
    !is.na(housing_cost_burden)
  ) %>%
  mutate(
    group       = paste(gender_head, "Mortgage Holder"),
    burden_pct  = housing_cost_burden * 100        # capped at 100%
  ) %>%
  group_by(multyear, group) %>%
  summarise(
    mean_cost_burden = weighted.mean(burden_pct, weight_household, na.rm = TRUE),
    .groups          = "drop"
  )

# position for annotations and compressed scale (same as before)
x_label <- min(data_gender_burden_mean$multyear) + diff(range(data_gender_burden_mean$multyear)) / 4
compress_trans <- trans_new(
  "compress",
  transform = function(x) ifelse(x <= 50, x, 50 + (x - 50) / 10),
  inverse   = function(y) ifelse(y <= 50, y, (y - 50) * 10 + 50)
)

# plot weighted‐mean burden
ggplot(data_gender_burden_mean,
       aes(multyear, mean_cost_burden, colour = group, linetype = group)) +
  annotate("rect", xmin = 2018, xmax = 2023, ymin = -Inf, ymax = Inf,
           fill = "lightyellow", alpha = 0.1) +
  geom_vline(xintercept = 2020, linetype = "dotted", colour = "lightgray", size = 1) +
  geom_hline(yintercept = c(30, 50, 100), colour = c("green", "orange", "red")) +
  annotate("text", x = x_label, y = c(30, 50, 100),
           label  = c("Affordable (<30%)","Moderate (30–50%)","Severe (>50%)"),
           colour = c("green","orange","red"), vjust = -0.4) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_y_continuous(
    trans  = compress_trans,
    name   = "Average Housing Cost Burden (%)",
    limits = c(0, NA)
  ) +
  scale_colour_manual(values = c(
    "Female Mortgage Holder" = "purple",
    "Male Mortgage Holder"   = "orange"
  )) +
  scale_linetype_manual(values = c(
    "Female Mortgage Holder" = "solid",
    "Male Mortgage Holder"   = "solid"
  )) +
  labs(
    title    = "Mean Housing Cost Burden for Mortgage Holders by Gender (2014–2023)",
    x        = "Year",
    colour   = "Group",
    linetype = "Group"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```
### A. Summary Table for Cost Burden (Owners) (Lara. Updated )
```{r}
data %>%
  filter(
    tenure_detailed == "Owned - With Mortgage or Loan",
    !is.na(housing_cost_burden)
  ) %>%
  group_by(multyear) %>%
  summarise(
    n             = n(),
    weighted_n    = sum(weight_household, na.rm=TRUE),
    mean_burden   = weighted.mean(housing_cost_burden, weight_household, na.rm=TRUE),
    median_burden = wtd.quantile(housing_cost_burden, weight_household, probs=0.5),
    p90_burden    = quantile(housing_cost_burden, 0.9, na.rm=TRUE),
    max_burden    = max(housing_cost_burden, na.rm=TRUE),
    .groups       = "drop"
  ) %>%
  print(n=Inf)
```

## VI. Ownership by Age and Race

### A. Age Ownership Analysis (Meg. RR TO REVISE )
```{r age-cleaning-ownership}
# Convert 90+ to numeric and clean
data$age_head[data$age_head == "90 (90+ in 1980 and 1990)"] <- "90"

#make age_head numeric
data$age_head <- as.numeric(data$age_head)
# check
summary(data$age_head)

#drop NAs
data_clean <- data%>% filter(!is.na(age_head))
#note: dropped about 12,000 NA's here 

# Clean and recode tenure variable
data_clean <- data_clean %>%
  filter(multyear >= 2014 & multyear <= 2024,
         !is.na(tenure_general),
         tenure_general != "Other/Unknown") %>%
  mutate(tenure_general = recode(tenure_general, "Owned" = "Owner", "Rented" = "Renter"),
         # ownership dummy
         owner = if_else(tenure_general == "Owner", 1, 0))

# check table
prop.table(table(data_clean$owner))

```
### B. Ownership Rate by multyear (Meg. RR)
```{r}
# ownership rate by multyear
ownership_by_multyear <- data_clean %>%
  group_by(multyear) %>%
  summarise(ownership_rate = mean(owner, na.rm = TRUE))

ggplot(ownership_by_multyear, aes(x = multyear, y = ownership_rate)) +
  annotate("rect", xmin = 2018, xmax = 2023, ymin = -Inf, ymax = Inf, fill = "yellow", alpha = 0.1) +
geom_vline(xintercept = 2020, linetype = "dotted", color = "lightgrey", linewidth = 1) +
  geom_line(color = "#3b7EA1", size = 1.5) +
  geom_point(color = "#F4A261", size = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     expand = expansion(mult = c(1, 1))) +  # <- this adds 5% padding at top/bottom
  labs(x = "multyear", y = "Total Ownership Rate", title = "CA Homeownership Rate 2014-2023") +
  theme_minimal()

#detailed ownership rates by multyear
tenure_prop_table <- data_clean %>%
  group_by(multyear, tenure_detailed) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(multyear) %>%
  mutate(proportion = count / sum(count)) %>%
  select(multyear, tenure_detailed, proportion) %>%
  pivot_wider(names_from = multyear, values_from = proportion, values_fill = 0)
print(tenure_prop_table) 
#why is ownership so high

sum(data_clean$owner, na.rm=TRUE)

```

### C. Ownership Rate by Age and multyear (Meg. RR)
```{r}
#summarize ownership by age group and multyear
data_ownership <- data_clean %>%
  group_by(multyear, age_head_group)%>%
  summarise(
    ownership_rate = mean(owner, na.rm = TRUE),
    count = n(),
    .groups = "drop" )

# Fix group_by in weighted summary
data_ownership_weighted <- data_clean %>%
  filter(!is.na(owner), !is.na(weight_household), !is.na(age_head_group), age_head > 17) %>%
  group_by(multyear, age_head_group) %>%  # <- USE age_head_group
  summarise(
    ownership_rate = sum(owner * weight_household) / sum(weight_household),
    count = n(),
    .groups = "drop"
  )

# Fix plot aesthetics
plot_share_owner <- ggplot(data_ownership_weighted, aes(x = multyear, y = ownership_rate, color = age_head_group, group = age_head_group)) +
  annotate("rect", xmin = 2018, xmax = 2023, ymin = -Inf, ymax = Inf, fill = "yellow", alpha = 0.1) +
  geom_vline(xintercept = 2020, linetype = "dotted", color = "lightgrey", linewidth = 1) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_text(aes(label = scales::percent(ownership_rate, accuracy = 1)), 
            vjust = -0.5, size = 3, show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Homeownership Rates by Age Cohort (2014–2023)",
    x = "Year",
    y = "Ownership Rate",
    color = "Age Cohort"
  ) +
  scale_color_manual(values = c(  "#D9661f", "#3b7EA1", "#859438", "#EE1f60", "#9467bd")) +
  theme_minimal() +
  theme(legend.position = "bottom")


print(plot_share_owner)

```

### D. Change in Ownership Rate  (Meg. RR)

```{r}
#show change in ownership rate (since first multyear. should we use a different starting place?)
data_ownership_share <- data_ownership_weighted %>%
  group_by(age_head_group) %>%
  arrange(multyear) %>%
  mutate(change_from_start = ownership_rate - first(ownership_rate))

plot_change_in_owner <- ggplot(data_ownership_share, aes(x = multyear, y = change_from_start, color = age_head_group, group = age_head_group)) + annotate("rect", xmin = 2018, xmax = 2023, ymin = -Inf, ymax = Inf, fill = "yellow", alpha = 0.1) +
geom_vline(xintercept = 2020, linetype = "dotted", color = "lightgrey", linewidth = 1) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Change in Homeownership Rate by Age Cohort (vs. Starting multyear)",
    x = "multyear",
    y = "Change from Starting Ownership Rate",
    color = "Age Cohort"
  ) +
   scale_color_manual(values = c("#D9661f", "#3b7EA1", "#859438", "#EE1f60"))  +
  theme_minimal() +
  theme(legend.position = "bottom")

ggsave("change_in_ownership_rate.png", plot = plot_change_in_owner, width = 8, height = 6, dpi = 300)
```

### E. Average Age of Owners vs Renter (Weighted)  (Meg. RR.)
```{r}
# Check :confirm that renters have valid age and weight value
data_clean %>%
  filter(tenure_general == "Renter") %>%
  summarise(
    total_rows = n(),
    missing_age = sum(is.na(age_head)),
    missing_weight = sum(is.na(weight_household)),
    nonmissing_both = sum(!is.na(age_head) & !is.na(weight_household))
  )

renter_summary <- data_clean %>%
  filter(tenure_general == "Renter", !is.na(age_head), !is.na(weight_household)) %>%
  group_by(multyear) %>%
  summarise(
    avg_age = sum(age_head * weight_household) / sum(weight_household),
    count = n(),
    .groups = "drop"
  )

print(renter_summary)

```

```{r}
# Clean and ensure tenure_general is a proper character/factor
data_clean <- data_clean %>%
  mutate(tenure_general = as_factor(tenure_general))

# Summarize with weighted average
data_age_summary_weighted <- data_clean %>%
  filter(!is.na(age_head), !is.na(weight_household), !is.na(tenure_general)) %>%
  group_by(multyear, tenure_general) %>%
  summarise(
    avg_age = sum(age_head * weight_household) / sum(weight_household),
    count = n(),
    .groups = "drop"
  )


plot_age <- ggplot(data_age_summary_weighted, aes(x = multyear, y = avg_age, color = tenure_general, group = tenure_general)) +
  annotate("rect", xmin = 2018, xmax = 2023, ymin = -Inf, ymax = Inf, fill = "yellow", alpha = 0.1) +
  geom_vline(xintercept = 2020, linetype = "dotted", color = "lightgrey", linewidth = 1) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Average Age of Renters vs. Owners (2016–2023)",
    x = "Year",
    y = "Average Age of Head of Household",
    color = "Tenure Type"
  ) +
  scale_color_manual(values = c("Owner" = "#EE1f60", "Renter" = "#3b7EA1")) +
  scale_y_continuous(limits = c(30, 60)) +
  theme_minimal() +
  theme(legend.position = "bottom")
print(plot_age)

ggsave("avg_age_renters_vs_owners.png", plot = plot_age, width = 8, height = 6, dpi = 300)

```


## VII. Homeownership and Cost Burden by Race

### A. Weighted Ownership by Race (Meg)
```{r ownership-by-race}
data_summary_race_weighted <- data %>%
  filter(multyear >= 2014 & multyear <= 2023,
         !is.na(tenure_general),
         !is.na(race_ethnicity_head),
         !is.na(weight_household)) %>%
  mutate(owner = if_else(tenure_general == "Owned", 1, 0)) %>%
  group_by(multyear, race_ethnicity_head) %>%
  summarise(
    ownership_rate = sum(owner * weight_household) / sum(weight_household),
    count = n(),
    .groups = "drop"
  )

ggplot(data_summary_race_weighted, aes(x = multyear, y = ownership_rate, color = race_ethnicity_head)) +
  annotate("rect", xmin = 2018, xmax = 2023, ymin = -Inf, ymax = Inf, fill = "yellow", alpha = 0.1) +
  geom_vline(xintercept = 2020, linetype = "dotted", color = "lightgrey", linewidth = 1) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "CA Homeownership Rates by Race (2014–2023)",
       x = "multyear", y = "Ownership Rate", color = "Race") +
  theme_minimal() +
  theme(legend.position = "right")
```

## VIII. Housing Burden

```{r }
 households <- data %>%
  mutate(
    income = as.numeric(income),
    weight_household = as.numeric(weight_household),
    puma = as.integer(puma),
    multyear = as.integer(multyear)
  )
```

### A. AMI  (KP)
```{r}
ami_by_puma_multyear <- households %>%
  group_by(puma, multyear) %>%
  summarise(AMI = weightedMedian(income, w = weight_household, na.rm = TRUE), .groups = "drop")
ami_by_puma_multyear 
```

#### 0. Define AMI Category
```{r ami-categorization}
households <- households %>%
  left_join(ami_by_puma_multyear, by = c("puma", "multyear")) %>%
  mutate(
    percent_ami = income / AMI,
    ami_category = case_when(
      percent_ami < 0.15 ~ "0–15% AMI",
      percent_ami >= 0.15 & percent_ami < 0.30 ~ "15–30% AMI",
      percent_ami >= 0.30 & percent_ami < 0.50 ~ "30–50% AMI",
      percent_ami >= 0.50 & percent_ami < 0.80 ~ "50–80% AMI",
      percent_ami >= 0.80 & percent_ami < 1.20 ~ "80–120% AMI",
      percent_ami >= 1.20 ~ "120%+ AMI",
      TRUE ~ NA_character_
    ),
    mi_category = factor(
      ami_category,
      levels = c("0–15% AMI", "15–30% AMI", "30–50% AMI", "50–80% AMI", "80–120% AMI", "120%+ AMI")
    )
  )

```

-----
#### 0. (REVISE) :Clean summary table by AMI category and year
```{r}
burden_summary <- households %>%
  filter(!is.na(ami_category), !is.na(cost_burden_category), multyear >= 2018) %>%
  mutate(
    overburdened = housing_cost_burden > 0.3,
    severely_burdened = cost_burden_category == "Severe (>50%)"
  ) %>%
  group_by(multyear, ami_category) %>%
  summarise(
    overburdened_rate = weighted.mean(overburdened, weight_household, na.rm = TRUE),
    severe_rate = weighted.mean(severely_burdened, weight_household, na.rm = TRUE),
    .groups = "drop"
  )


burden_summary %>%
  arrange(multyear, ami_category) %>%
  kable(digits = 3, format = "html", caption = "Housing Cost Burden Rates by AMI and Year") %>%
  kable_styling(full_width = FALSE, position = "left")

```

#2. Bar Plot: Overburdened Share by AMI Category
```{r}
households %>%
  filter(!is.na(ami_category), !is.na(housing_cost_burden)) %>%
  mutate(overburdened = housing_cost_burden > 0.3) %>%
  group_by(ami_category) %>%
  summarise(rate = weighted.mean(overburdened, weight_household, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(ami_category, -rate), y = rate)) +
  geom_col(fill = "#3b7ea1") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Share of Overburdened Households by AMI Category",
       x = "AMI Category", y = "Overburdened Share") +
  coord_flip() +
  theme_minimal()

```

#3. Stacked Bar Plot: Burden Category Breakdown by AMI and Year

```{r}
plot_data <- households %>%
  filter(!is.na(ami_category), !is.na(cost_burden_category), multyear >= 2018) %>%
  group_by(multyear, ami_category, cost_burden_category) %>%
  summarise(n = sum(weight_household, na.rm = TRUE), .groups = "drop") %>%
  group_by(multyear, ami_category) %>%
  mutate(share = n / sum(n))

ggplot(plot_data, aes(x = ami_category, y = share, fill = cost_burden_category)) +
  geom_col(position = "stack") +
  facet_wrap(~ multyear, nrow = 1) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Burden Level Breakdown by AMI Category and Year",
       x = "AMI Category", y = "Share of Households", fill = "Burden Level") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")

```
-----
#### 1. Trend Over Time by AMI Category
Shows how cost burden changed over time within each AMI group.
```{r plot-burden-by-ami}
households %>%
  filter(multyear >= 2018, !is.na(ami_category), !is.na(housing_cost_burden)) %>%
  mutate(burden_flag = housing_cost_burden > 0.3) %>%
  group_by(multyear, ami_category) %>%
  summarise(overburdened = weighted.mean(burden_flag, weight_household, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = multyear, y = overburdened, color = ami_category)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Cost Burden Over Time by AMI Category (2018–2023)",
    x = "Year", y = "Percent Overburdened", color = "AMI Category"
  ) +
  theme_minimal()

```
#### 2. Boxplot of Cost Burden Distribution
Shows how much variation there is in cost burden within each AMI level.
```{r}
households %>%
  filter(!is.na(ami_category), !is.na(housing_cost_burden)) %>%
  ggplot(aes(x = ami_category, y = housing_cost_burden)) +
  geom_boxplot(fill = "#3b7ea1", alpha = 0.7) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Distribution of Housing Cost Burden by AMI Category",
    x = "AMI Category", y = "Housing Cost Burden (as % of Income)"
  ) +
  theme_minimal()

```
#### 3. Population Share Annotations
Still want to show the average overburden rate by AMI, but now with context on group size.
```{r}
households %>%
  filter(!is.na(ami_category), !is.na(housing_cost_burden)) %>%
  mutate(burden_flag = housing_cost_burden > 0.3) %>%
  group_by(ami_category) %>%
  summarise(
    n = sum(weight_household, na.rm = TRUE),
    overburdened = weighted.mean(burden_flag, weight_household, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("n = ", format(round(n), big.mark = ","))) %>%
  ggplot(aes(x = reorder(ami_category, -overburdened), y = overburdened)) +
  geom_col(fill = "#F4A261") +
  geom_text(aes(label = label), hjust = -0.1, size = 3.2) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = "Overburden Rate by AMI Category (with Population Size)",
    x = "AMI Category", y = "Percent Overburdened"
  ) +
  theme_minimal()
```
#### 4. Burden by AMI Category (selected years & bar)
```{r ami-cost-burden-stacked}
plot_data <- households %>%
  filter(multyear %in% c(2018, 2021, 2023),
         !is.na(ami_category),
         !is.na(cost_burden_category)) %>%
  group_by(multyear, ami_category, cost_burden_category) %>%
  summarise(households = sum(weight_household, na.rm = TRUE), .groups = "drop") %>%
  group_by(multyear, ami_category) %>%
  mutate(share = households / sum(households)) %>%
  ungroup()

ggplot(plot_data, aes(x = ami_category, y = share, fill = cost_burden_category)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  facet_wrap(~ multyear, nrow = 1) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(
    values = c("Affordable (<30%)" = "#F8766D",
               "Moderate (30-50%)" = "#00BFC4",
               "Severe (>50%)" = "#C77CFF",
               "Missing" = "gray70"),
    drop = FALSE
  ) +
  labs(
    title = "Cost Burden Distribution by AMI Category (2018, 2021, 2023)",
    x = "AMI Category", y = "Share of Households", fill = "Burden Level"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )

```
####  5.severe burden rate over time
```{r}
severe_burden <- households %>%
  filter(cost_burden_category == "Severe (>50%)",
         !is.na(ami_category)) %>%
  group_by(multyear, ami_category) %>%
  summarise(share_severe = sum(weight_household, na.rm = TRUE) / 
              sum(weight_household[!is.na(cost_burden_category)]), .groups = "drop")

ggplot(severe_burden, aes(x = multyear, y = share_severe, color = ami_category)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Severe Cost Burden Rates Over Time by AMI Category",
    x = "Year", y = "Percent Severely Burdened", color = "AMI Category"
  ) +
  theme_minimal()

```

#### 5. Burden by Race 
````{r}
cost_burden_summary <- households %>%
  filter(!is.na(housing_cost_burden),
         !is.na(weight_household),
         !is.na(race_ethnicity_head),
         housing_cost_burden <= 1) %>%  # remove bad values
  mutate(burden_flag = housing_cost_burden > 0.3) %>%
  group_by(race_ethnicity_head) %>%
  summarise(
    overburdened = weighted.mean(burden_flag, weight_household, na.rm = TRUE),
    count = n(),
    .groups = "drop"
  )

ggplot(cost_burden_summary, aes(x = reorder(race_ethnicity_head, -overburdened), y = overburdened)) +
  geom_col(fill = "#3b7EA1") +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  labs(
    title = "Share of Households Overburdened by Housing Costs",
    y = "Percent Overburdened", x = "Race/Ethnicity"
  ) +
  theme_minimal()


````

#### 6. Cost Burden by Race and Year (2018–2023)
```{r cost-burden-by-race-time}
# Step 1: Ensure correct data types
households <- households %>%
  mutate(
    housing_cost_burden = as.numeric(housing_cost_burden),
    weight_household = as.numeric(weight_household)
  )

# Step 2: Compute overburdened flag and summarize by year and race
cost_burden_by_multyear_race <- households %>%
  filter(multyear >= 2018,
         !is.na(housing_cost_burden),
         !is.na(race_ethnicity_head),
         !is.na(weight_household)) %>%
  mutate(burden_flag = housing_cost_burden > 0.3) %>%
  group_by(multyear, race_ethnicity_head) %>%
  summarise(overburdened = weighted.mean(burden_flag, weight_household, na.rm = TRUE),
            .groups = "drop")

# Step 3: Plot with optional annotation for 2020
ggplot(cost_burden_by_multyear_race,
       aes(x = multyear, y = overburdened, color = race_ethnicity_head)) +
  annotate("rect", xmin = 2020, xmax = 2021, ymin = 0, ymax = 1,
           fill = "lightgray", alpha = 0.1) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "Cost Burden by Race and Year (2018–2023)",
       x = "Year",
       y = "Percent Overburdened",
       color = "Race/Ethnicity") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right")


```

## IX. Time Series: Ownership and Cost Burden Trends
```{r homeownership-trends}
home_by_multyear <- households %>%
  filter(!is.na(tenure_general), multyear >= 2018) %>%
  mutate(homeowner = tenure_general == "Owned") %>%
  group_by(multyear) %>%
  summarise(home_own_rate = weighted.mean(homeowner, weight_household))

ggplot(home_by_multyear, aes(x = multyear, y = home_own_rate)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Homeownership Rate in California (2018–2023)",
       y = "Homeownership Rate", x = "Year")
```

## X. Descriptive Ownership Snapshots by Demographics

### A. Overall Distribution
```{r overall-distribution}

overall_distribution <- dplyr::count(households, tenure_detailed, wt = weight_household) %>%
  mutate(share = n / sum(n)) %>%
  arrange(desc(share))

print(overall_distribution)

```

### B. By Race
```{r by-race}
households %>%
  group_by(race_ethnicity_head, tenure_detailed) %>%
  summarise(n = sum(weight_household, na.rm = TRUE), .groups = "drop") %>%
  group_by(race_ethnicity_head) %>%
  mutate(share = n / sum(n)) %>%
  ggplot(aes(x = tenure_detailed, y = share, fill = race_ethnicity_head)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Ownership by Race/Ethnicity", x = "Tenure Type", y = "Share") +
  theme_minimal() +
  theme(legend.position = "top")

```


### C. By Veteran and Disability Status
```{r veteran}
households %>%
  group_by(veteran_head, tenure_detailed) %>%
  summarise(n = sum(weight_household, na.rm = TRUE), .groups = "drop") %>%
  group_by(veteran_head) %>%
  mutate(share = n / sum(n)) %>%
  ggplot(aes(x = tenure_detailed, y = share, fill = veteran_head)) +
  geom_col(position = "dodge") +
  labs(title = "Ownership by Veteran Status", x = "Tenure", y = "Share") +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() +
  theme(legend.position = "top")


```

```{r Disability}
households %>%
  group_by(disability_head, tenure_detailed) %>%
  summarise(n = sum(weight_household, na.rm = TRUE), .groups = "drop") %>%
  group_by(disability_head) %>%
  mutate(share = n / sum(n)) %>%
  ggplot(aes(x = tenure_detailed, y = share, fill = disability_head)) +
  geom_col(position = "dodge") +
  labs(title = "Ownership by Disability Status", x = "Tenure", y = "Share") +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() +
  theme(legend.position = "top")


```


### D. By Age Group
```{r by-age}
households %>%
  filter(!is.na(age_head_group)) %>%
  group_by(age_head_group, tenure_detailed) %>%
  summarise(n = sum(weight_household, na.rm = TRUE), .groups = "drop") %>%
  group_by(age_head_group) %>%
  mutate(share = n / sum(n)) %>%
  ggplot(aes(x = tenure_detailed, y = share, fill = age_head_group)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Ownership by Age Group of Household Head",
    x = "Tenure Type", y = "Share",
    fill = "Age Group"
  ) +
  theme_minimal() +
  theme(legend.position = "top")


```


